---
title: "01-5_check_name"
subtitle: "Check name variables"
author: "Ross Gayler"
date: "2021-01-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup}
# Set up the project environment, because each Rmd file knits in a new R session
# so doesn't get the project setup from .Rprofile

# Project setup
library(here)
source(here::here("code", "setup_project.R"))

# Extra set up for the 01*.Rmd notebooks
source(here::here("code", "setup_01.R"))

# Extra set up for this notebook
# ???

# start the execution time clock
tictoc::tic("Computation time (excl. render)")
```

# Introduction

The `01*.Rmd` notebooks read the data, filter it to the subset to be
used for modelling, characterise it to understand it, check for possible
gotchas, clean it, and save it for the analyses proper.

This notebook (`01-5_check_name`) characterises the name variables in
the saved subset of the data.

These variables will be used to construct the main predictors in the
compatibility models.

We intend to use the one snapshot file as both the database to be
queried and as the set of queries. Consequently, strictly speaking, we
don't need to standardise the name variables because the database and
query records are guaranteed to be identical (they will literally be the
same record). However, we will look at the name variables with an eye to
standardisation because it is never a good idea to statistically model
data without having an idea about the quality of the data. We will apply
some basic standardisation to the name variables, if appropriate,
because it parallels what would be necessary in practice.

------------------------------------------------------------------------

Define the name variables.

```{r}
vars_name <- c(
  "last_name", "first_name", "midl_name", "name_sufx_cd" 
)
```

Read the usable data. Remember that this consists of only the ACTIVE &
VERIFIED records.

```{r}
# Show the entity data file location
# This is set in code/file_paths.R
fs::path_file(f_entity_raw_fst)

# get data for next section of analyses
d <- fst::read_fst(
  f_entity_raw_fst, 
  columns = c(vars_name, "sex") # get sex as well for cross-checking
) %>% 
  tibble::as_tibble()
dim(d)
```

Take a quick look at the distributions.

```{r}
d %>% skimr::skim()
```

-   `last_name` 100% filled
-   `first_name` \~100% filled (23 missing)
-   `midl_name` 94% filled
-   `name_sufx_cd` 6% filled

# Name length

Look at the distributions of name lengths first, before moving on to
analyses more focused on standardisation.

Calculate the lengths of the name variables.

```{r}
x <- d %>% 
  dplyr::mutate(
    len_last = stringr::str_length(last_name),
    len_first = stringr::str_length(first_name),
    len_midl = stringr::str_length(midl_name)
  )
```

## last_name

`last_name` Voter last name

Look at the distributions of name lengths.

```{r}
summary(x$len_last)
table(x$len_last, useNA = "ifany")

x %>% 
  ggplot() +
  geom_histogram(aes(x = len_last), binwidth = 1) +
  scale_y_sqrt()
```

Look at examples of short names.

```{r}
# length == 1
x %>% 
  dplyr::filter(len_last == 1) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

-   1-letter last names are very rare
-   1-letter last names are probably errors

```{r}
# length == 2
x %>% 
  dplyr::filter(len_last == 2) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

-   Most 2-letter last names are probably valid.
-   ST is probably Saint from a multi-word last name

Look at examples of long names.

```{r}
# length == 21
x %>% 
  dplyr::filter(len_last == 21) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

-   21-letter last names are hyphenated

```{r}
# length >= 20
x %>% 
  dplyr::filter(len_last >= 20) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

-   20+-letter last names appear to be multi-word and/or hyphenated

## first_name

`first_name` Voter first name

Look at the distributions of name lengths.

```{r}
summary(x$len_first)
table(x$len_first, useNA = "ifany")

x %>% 
  ggplot() +
  geom_histogram(aes(x = len_first), binwidth = 1) +
  scale_y_sqrt()
```

Look at the missing names.

```{r}
x %>% 
  dplyr::filter(is.na(first_name)) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

-   Some missing first names look like the middle name is actually the
    first name, e.g. ? JASON ALEXANDER
-   Some missing first names appear to have only a last name, e.g. ? ?
    AMEN
-   Some missing first names appear to have the entire name in the last
    name variable, e.g. ? ? FRYE WILLIAM C

Look at examples of short names.

```{r}
# length == 1
x %>% 
  dplyr::filter(len_first == 1) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

-   The 1-letter first names appear to be using an initial as the first
    name

```{r}
# length == 2
x %>% 
  dplyr::filter(len_first == 2) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

2-letter first names appear to be:

-   Valid, e.g. JO W CLARK, HO NGOC NGUYEN
-   Part of a multi word name that has bee split across the first and
    middle name variables, e.g. LA SONDA FOWLER

Look at the long names.

```{r}
# length >= 16
x %>% 
  dplyr::filter(len_first >= 16) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

Long first names appear to be:

-   Long non-anglo names, e.g. LAKSHMINARAYANAN
-   Multi-word and/or hyphenated, e.g. ELIZABETH-LINDSAY

## midl_name

`midl_name` Voter middle name

These names will often be missing or initials only.

Look at the distributions of name lengths.

```{r}
summary(x$len_midl)
table(x$len_midl, useNA = "ifany")

x %>% 
  ggplot() +
  geom_histogram(aes(x = len_midl), binwidth = 1) +
  scale_y_sqrt()
```

-   *Many* records are missing middle name
-   Spike of 1-letter names will be initials

Look at the long names.

```{r}
# lentgh >= 16
x %>% 
  dplyr::filter(len_midl >= 16) %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  knitr::kable()
```

-   Long middle names appear to be multiple names and/or hyphenated

```{r}
# clean up
rm(x)
gc()
```

# name_sufx_cd

`name_sufx_cd` Voter name suffix

This is intended for generation markers, e.g. Junior, Senior.

I am not going to use name suffix in entity resolution because age
should be sufficient and is much better quality. I will look at what
values turn up in the name suffix because the same values sometimes
wrongly occur in the main name variables. Knowing what values occur may
help us to remove those values from the main name variables.

```{r}
d %>% dplyr::select(name_sufx_cd) %>% skimr::skim()
table(d$name_sufx_cd, useNA = "ifany") %>% sort() %>% rev()

# get a better look at the cleaned suffixes
d %>% 
  dplyr::mutate(
    sufx = name_sufx_cd %>% 
      stringr::str_to_upper() %>% 
      stringr::str_remove_all(pattern = "[^A-Z0-9]") %>% # remove non-alphanumeric
      dplyr::na_if("") 
  ) %>% 
  dplyr::count(sufx) %>% 
  dplyr::filter(n > 1) %>% 
  dplyr::arrange(desc(n), sufx) %>% 
  knitr::kable()
```

-   There are generation suffixes: JR, SR, I, II (11), III (111), IV, V,
    VI, VII
-   There are honorific titles: MRS, MR, MS, DR, REV

# Standardisation

Look at issues that might be addressed by standardisation.

For each type of standardisation issue look at first middle and last
names separately, because the issue may manifest differently in each of
the name variables.

## Lower-case letters.

```{r}
d %>% dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "[a-z]"))

d %>% dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "[a-z]"))

d %>% dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "[a-z]"))
```

-   Lower case letters occur in last, first, and middle names
-   Associated with particles where there would optionally be a space,
    e.g. JoANN, McBride

## Non-alphanumeric

Check for non-alphanumeric characters in names.

### Hyphen

Check for hyphens.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "-"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, sex) %>% 
  knitr::kable()
```

-   \~21k last names with hyphens
-   Look like legitimately hyphenated last names

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "-"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name, sex) %>% 
  knitr::kable()
```

-   \~3kL first names with hyphens
-   Look like legitimately hyphenated first names

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "-"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name, sex) %>% 
  knitr::kable()
```

-   ~4k middle names with hyphens
-   Look like legitimately hyphenated middle names

### Quote

Check for quotes.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "'"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, sex) %>% 
  knitr::kable()
```

-   ~5k last names with quotes
-   Look like legitimately quoted last names

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "'"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name, sex) %>% 
  knitr::kable()
```

-   ~1k first names with quotes
-   Look like legitimately quoted first names

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "'"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name, sex) %>% 
  knitr::kable()
```

-   ~3k middle names with quotes
-   Look like legitimately quoted middle names

### Period

Check for periods.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "\\."))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, sex) %>% 
  knitr::kable()
```

-   11 last names with periods
-   Look like legitimate abbreviations

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "\\."))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name, sex) %>% 
  knitr::kable()
```

-   120 first names with periods
-   Look like initials

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "\\."))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name, sex) %>% 
  knitr::kable()
```

-   ~2k middle names with periods
-   Look like initials

### Comma

Check for commas.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, ","))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, sex) %>% 
  knitr::kable()
```

-   2 last names with commas
-   Punctuation for suffix field values added to last name

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, ","))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name, sex) %>% 
  knitr::kable()
```

-   4 first names with commas
-   Arbitrary added punctuation
-   Punctuation for suffix field value added to first name

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, ","))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name, sex) %>% 
  knitr::kable()
```

-   12 middle names with periods
-   List separator
-   Punctuation to squeeze in extra field

### Other non-alphanumeric

Check for other non-alphanumeric characters.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "[^ a-zA-Z0-9\\.,'-]"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, sex) # %>% 
  # knitr::kable() # some of the characters break the kable formatting
```

-   31 last names with other non-alphanumeric characters
-   Most look like substitutions for hyphen or quote
-   Some look like random cruft

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "[^ a-zA-Z0-9\\.,'-]"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name, sex) # %>% 
  # knitr::kable() # some of the characters break the kable formatting
```

-   102 first names with other non-alphanumeric characters
-   Some look like substitutions for hyphen or quote
-   Some are parenthetical notes
-   Some look like random cruft

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "[^ a-zA-Z0-9\\.,'-]"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name, sex) %>% 
  knitr::kable()
```

-   ~1k middle names with other non-alphanumeric characters
-   Some look like substitutions for hyphen
-   Many are parenthetical notes (NMN = no middle name)

## Digits

Check for digits.

### Zero

Check for zero

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "0"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, sex) %>% 
  knitr::kable()
```

-   29 last names with zero
-   Substitution for O

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "0"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>%
  dplyr::arrange(first_name, sex) %>% 
  knitr::kable()
```

-   33 first names with zero
-   Substitution for O

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "0"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>%
  dplyr::arrange(midl_name, sex) %>% 
  knitr::kable()
```

-   77 middle names with zero
-   Some are substitution for O
-   Some are in superfluous numbers

### One

Check for one.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "1"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, sex) %>% 
  knitr::kable()
```

-   1 last name with one
-   Substitution for I in generation suffix (111 = III)

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "1"))

nrow(x)
```

-   0 first names with one

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "1"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>%
  dplyr::arrange(midl_name, sex) %>% 
  knitr::kable()
```

-   39 middle names with one
-   Some are substitution for I in generation suffix
-   Some are in superfluous numbers

### Other digits

Check for other digits.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "[2-9]"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name, sex) %>% 
  knitr::kable()
```

-   1 last name with a 5
-   Random insertion

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "[2-9]"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>%
  dplyr::arrange(first_name, sex) %>% 
  knitr::kable()
```

-   2 first names with digits 2-9
-   Look like random insertions

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "[2-9]"))

nrow(x)

x %>%   
  dplyr::slice_sample(n = 20) %>%
  dplyr::arrange(midl_name, sex) %>% 
  knitr::kable()
```

-   24 middle names with digits 2-9
-   One random insertion
-   Most appear to be superfluous numbers (from the address?)

## Special words

Look for special words that shouldn't be in names.

Define word patterns to search for.

```{r}
# honorifics
w_hons <- c(
  "MR", "MISTER", "MASTER", "MRS", "MS", "MISS", 
  "REV", "REVEREND", "SR", "SISTER", "BR", "BROTHER",
  "FATHER", "MOTHER", "PASTOR", "ELDER", "BISHOP",
  "DR", "DOCTOR", "MD", "PROF", "PROFESSOR"
)

# generation suffixes
w_gen <- c(
  "JR", "JNR", "JUNIOR", "SR", "SNR", "SENIOR",
  "1ST", "2ND", "3RD", "4TH", "5TH", "6TH", "7TH", "8TH",
  "FIRST", "SECOND", "THIRD", "FOURTH", "FIFTH", "SIXTH", "SEVENTH", "EIGHTH", "EIGHTTH",
  "1", "2", "3", "4", "5", "6", "7", "8",
  "I", "II", "III", "IIII", "IV", "V", "VI"
)

# special values
w_spec <- c(
  "NN", "NMN", "NAME",
  "UNK", "UNKNOWN", "AKA", "KNOWN AS", "ALSO KNOWN AS", "ALIAS",
  "BLIND"
)

# test
w_test <- c(
  "TEST", "TST", "DUMMY", "VOTER",  "([A-Z])\\1{2,}"
)
```

### Last name

```{r}
# regular expression to match words
w_regexp <- 
  c(w_hons, w_gen, w_spec, w_test) %>% # all special words
  unique() %>% # make it a set
  dplyr::setdiff( # remove words that appear to mostly be validly used
    c(
      "BISHOP",
      "BLIND",
      "BROTHER",
      "DOCTOR",
      "ELDER",
      "FIRST",
      "JUNIOR",
      "MASTER",
      "MISS",
      "MISTER",
      "PASTOR",
      "SENIOR",
      "TEST",
      "THIRD",
      "VOTER"
    )
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") # search for any

x <- d %>% 
  dplyr::mutate(
    match = 
      last_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_extract(pattern = w_regexp)
  ) %>% 
  dplyr::filter(!is.na(match))

nrow(x)

x %>% 
  dplyr::arrange(match, sex, last_name, first_name) %>% 
  knitr::kable()
```

I eyeballed the results and removed words which appeared to be mostly
validly used.

Invalid words:

-   As whole field:
-   As first word:
-   As last word: DR, II, III, IIII, IV, JR, MD, SR
-   As internal word: SR

### First name

```{r}
# regular expression to match words
w_regexp <- 
  c(w_hons, w_gen, w_spec, w_test) %>% # all special words
  unique() %>% # make it a set
  dplyr::setdiff( # remove words that appear to mostly be validly used
    c(
      "BISHOP",
      "BROTHER",
      "DOCTOR",
      "ELDER",
      "JUNIOR",
      "MASTER",
      "MISTER",
      "PASTOR",
      "PROFESSOR"
    )
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") # search for any

x <- d %>% 
  dplyr::mutate(
    match = 
      first_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_extract(pattern = w_regexp)
  ) %>% 
  dplyr::filter(!is.na(match))

nrow(x)

x %>% 
  dplyr::arrange(match, sex, last_name, first_name) %>% 
  knitr::kable()
```

I eyeballed the results and removed words which appeared to be mostly
validly used.

Invalid words:

-   As whole field: FATHER, III, IV, JR, MD, MR, MRS, SISTER, SR
-   As first word: DR, MISS, MRS, REV, SISTER
-   As last word: III, JR, MRS, NMN, SR
-   As internal word: MRS

### Middle name

```{r}
# regular expression to match words
w_regexp <- 
  c(w_hons, w_gen, w_spec, w_test) %>% # all special words
  unique() %>% # make it a set
  dplyr::setdiff( # remove words that appear to mostly be validly used
    c(
      "BISHOP",
      "BLIND",
      "BR",
      "BROTHER",
      "DOCTOR",
      "ELDER",
      "FIRST",
      "JR", # invalid & too many to display 
      "JUNIOR",
      "MASTER",
      "MISTER",
      "MRS", # invalid & too many to display
      "NMN", # invalid & too many to display
      "PASTOR",
      "SENIOR",
      "SISTER",
      "I",
      "V",
      "VI",
      "VOTER"
    )
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") # search for any

x <- d %>% 
  dplyr::mutate(
    match = 
      midl_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_extract(pattern = w_regexp)
  ) %>% 
  dplyr::filter(!is.na(match))

nrow(x)

x %>% 
  dplyr::arrange(match, sex, last_name, first_name) %>% 
  knitr::kable()
```

I eyeballed the results and removed words which appeared to be mostly
validly used.

Invalid words:

-   As whole field: AKA, DR, II, III, IV, JR, MD, MISS, MRS, MS, NMN,
    REV, SR
-   As first word: JR, MRS
-   As last word: DR, II, III, IV, JR, MD, MISS, MR, MRS, NMN, NN, SR
-   As internal word: JR

# Timing {.unnumbered}

```{r echo=FALSE}
tictoc::toc()
```
