---
title: "01_get_check_data"
author: "Ross Gayler"
date: "2020-12-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup}
library(here)
library(magrittr)
library(dplyr)
library(stringr)
library(vroom)
library(skimr)
library(knitr)
library(glue)
```

# Introduction

Read the data, characterise it to understand it, and check for possible
gotchas.

This project uses historical voter registration data from the [North
Carolina State Board of Elections](https://www.ncsbe.gov/). This
information is made publicly available in accordance with [North
Carolina state
law](https://s3.amazonaws.com/dl.ncsbe.gov/ReadMe_PUBLIC_DATA.txt). The
[Voter Registration Data
page](https://www.ncsbe.gov/results-data/voter-registration-data) links
to a [folder of Voter Registration
snapshots](https://dl.ncsbe.gov/index.html?prefix=data/Snapshots/),
which contains the snapshot data files and a [metadata file describing
the layout of the snapshot data
files](https://s3.amazonaws.com/dl.ncsbe.gov/data/Snapshots/layout_VR_Snapshot.txt).
At the time of writing the snapshot files cover the years 2005 to 2020
with at least one snapshot per year. The files are [ZIP
compressed](https://en.wikipedia.org/wiki/ZIP_(file_format)) and
relatively large, with the smallest being 572 MB after compression.

The snapshots contain many columns that are irrelevant to this project
and/or prohibited under Australian privacy law (e.g. political
affiliation, race). We initially read *all* the columns, because that
may help debugging the inevitable problems reading the data. Later the
data set will be restricted to the essential columns for the project.

We use only one snapshot file
([VR_Snapshot_20051125.zip](https://s3.amazonaws.com/dl.ncsbe.gov/data/Snapshots/VR_Snapshot_20051125.zip))
because this project does not investigate linkage of records across
time. We chose the oldest snapshot (2005) because it is the smallest and
the contents are the most out of date, minimising the current
information made available. Note that this project will not generate any
information that is not already directly, publicly available from NCSBE.

We intend to use the one snapshot file as both the database to be
queried and as the set of queries. Consequently, strictly speaking, we
don't need to standardise the name attributes because the database and
query records are guaranteed to be identical (they will literally be the
same record). However, we will look at the attributes with an eye to
standardisation because it is never a good idea to statistically model
data without having an idea about the quality of the data. We will apply
some basic standardisation to the attributes if appropriate because it
parallels what would be necessary in practice.

# Read data

The snapshot ZIP file was downloaded, uncompressed (5.7 GB), then
compressed in [XZ format](https://en.wikipedia.org/wiki/XZ_Utils) to
minimise the size. The compressed snapshot file and the metadata file
are stored in the `data` directory.

```{r}
raw_file <- here::here("data", "VR_20051125.txt.xz") # raw input file
```

The cleaned data is stored as an [`fst`
format](https://www.fstpackage.org/) file in the `output` directory.

```{r}
d_fst <- here::here("output", "d.fst") # temporary data file
clean_fst <- here::here("output", "clean.fst") # parsed and cleaned data as a dataframe
```

The data is tab-separated, not fixed-width as you might reasonably think
from reading the metadata. The field widths (interpreted as maximum
lengths) in the metadata are not accurate. Some fields contain values
longer than the stated width.

Inspection of the raw data shows that the character fields are unquoted.
However, at least one character value contains a double-quote character,
which has the potential to confuse the parsing if it is looking for
quoted values.

```{r eval=FALSE}
d <- vroom::vroom( #read raw data; let vroom guess the field types
  raw_file,
  delim = "\t", # assume that fields are *only* delimited by tabs
  col_names = TRUE, # use the column names on the first line of data
  na = "", # missing fields are empty string or whitespace only (see trim_ws argument)
  quote = "", # don't allow for quoted strings
  comment = "", # don't allow for comments
  trim_ws = TRUE, # trim leading and trailing whitespace
  escape_double = FALSE, # assume no escaped quotes
  escape_backslash = FALSE # assume no escaped backslashes
  )
fst::write_fst(d, d_fst, compress = 100) # save data frame (cheap-skate caching)
```

Some of the analyses have been done on a laptop with 16GB of RAM. The
data set is almost too big for that laptop, so for different sections of
the analysis I read a subset of the columns from the temporary data
file, delete the dataframes after use and clean up the RAM with a
garbage collection.

```{r}
d <- fst::read_fst(d_fst) %>% tibble::as_tibble() # get cached data
dim(d)
```

-   Correct number of data rows extracted (external line count of input
    file = 8,003,294)

# Characterise data (all records)

Take a very quick look at everything then concentrate on the columns
that have a chance of being useful.

```{r}
glimpse(d)
skimr::skim(d)
```

-   The warning messages from `skim()` indicate that a handful of rows
    contain unexpected characters. If they are in rows we use they will
    have to be located and dealt with.

```{r}
# clean up
rm(d)
gc()

# get data for next section of analyses
d <- fst::read_fst(
  d_fst,
  columns = c("county_id", "county_desc", "voter_reg_num", "ncid", "status_cd", 
              "voter_status_desc", "reason_cd", "voter_status_reason_desc")
  ) %>% 
  tibble::as_tibble()
dim(d)
```

## county_id & county_desc

`county_id`: County identification number\
`county_desc`: County description

```{r}
summary(d$county_id)
table(d$county_id)
```

-   Never missing
-   Integer 1 .. 100

```{r}
table(d$county_desc)
```

-   Never missing
-   100 unique values

They look reasonable, to the extent that I can tell without knowing
anything about the counties.

## voter_reg_num

`voter_reg_num`: Voter registration number (unique by county)

```{r}
table(d$voter_reg_num) %>% head(12)
table(d$voter_reg_num) %>% tail(12)

summary(as.integer(d$voter_reg_num))

d$voter_reg_num %>% stringr::str_length() %>% table(useNA = "ifany")
```

-   \~2.7M unique values
-   Never missing
-   Integer 0 .. \~1,000M (as strings)
-   Looks like they should be 12-digit integers with leading zeroes
-   Exactly one observation is short

Look at the record with the short value.

```{r}
d %>% 
  dplyr::filter(stringr::str_length(voter_reg_num) < 12) %>% 
  dplyr::select(county_id, voter_reg_num, status_cd, voter_status_desc, reason_cd, voter_status_reason_desc) %>% 
  knitr::kable()
```

-   There is only one short value which can be ignored because it will
    later be excluded from the data set because of the observation's
    status -not active. (I intend to later restrict the data set to only
    active voters because, to the greatest extent possible, I want to
    have no duplicate records in the data used for the analyses.)

Check whether `county_id x voter_reg_num` is unique, as claimed.

```{r}
d %>% 
  dplyr::select(county_id, voter_reg_num) %>% 
  dplyr::mutate(id = stringr::str_c(as.character(county_id), ".", voter_reg_num)) %>% 
  dplyr::count(id) %>% 
  with(table(n))
```

-   `county_id x voter_reg_num` is unique, even including observations
    flagged as duplicates.

## ncid

`ncid`: North Carolina identification number (NCID) of voter

-   Always missing

That's a shame. It would have been useful.

## status_cd & voter_status_desc

`status_cd`: Status code for voter registration\
`voter_status_desc`: Status code description

```{r}
table(d$status_cd, useNA = "always")
table(d$voter_status_desc, useNA = "always")
```

-   5 unique nonmissing values
-   2 records with missing values
-   \~4.9M active records

## reason_cd & voter_status_reason_desc

`reason_cd`: Reason code for voter registration status\
`voter_status_reason_desc`: Reason code description

```{r}
table(d$reason_cd, useNA = "always")
table(d$voter_status_reason_desc, useNA = "always")
```

-   26 unique nonmissing values
-   238 records with missing values
-   \~4.1M verified records

Look at the relationship between status and status reason.

```{r}
table(
  stringr::str_trunc(d$voter_status_reason_desc, 25), 
  stringr::str_trunc(d$voter_status_desc, 8), 
  useNA = "always"
)
```

-   voter_status_desc == "ACTIVE" & voter_status_reason_desc ==
    "VERIFIED"

    -   Most likely to be error free (based on common-sense
        interpretation of the labels)
    -   \~4.1M observations

## Name standardisation

Identify any oddities about the name fields that might benefit from
standardisation.

I will do this on all the rows, not just the subset to be analysed,
because I expect the oddities to be much the same independently of
whether I will exclude the rows from the analyses and the larger sample
size will be helpful in spotting rare problems.

I will look at the three name fields concurrently because I expect the
oddities to be similar across the name fields.

-   `last_name`: Voter last name
-   `first_name`: Voter first name
-   `midl_name`: Voter middle name

Look for possible anomalies in names.

```{r}
# clean up
rm(d)
gc()

# get data for next section of analyses
d <- fst::read_fst(
  d_fst,
  columns = c(
    "last_name", "first_name", "midl_name", "name_sufx_cd", 
    "sex", "age", "voter_status_desc", "voter_status_reason_desc"
  )
) %>% 
  tibble::as_tibble()
dim(d)
```

### Name missing

```{r}
d %>% with(table(is.na(last_name)))
d %>% with(table(is.na(first_name)))
d %>% with(table(is.na(midl_name)))
```

-   A small fraction of last and first names are missing. We don't
    expect them to be missing.
-   A significant fraction of middle names are missing. This is expected
    as middle names are not mandatory.

Look at the records missing last or first names to see if there is some
explanation for their absence.

```{r}
# last name missing
d %>% 
  dplyr::filter(is.na(last_name)) %>% 
  dplyr::select(
    first_name, midl_name, name_sufx_cd, 
    sex, age,
    voter_status_desc, voter_status_reason_desc
    ) %>% 
  dplyr::arrange(voter_status_desc, voter_status_reason_desc, first_name) %>% 
  knitr::kable()
```

-   All the voters missing `last_name` are REMOVED. Perhaps it's a
    side-effect of the removal process.

```{r}
# first name missing
d %>% 
  dplyr::filter(is.na(first_name)) %>% 
  dplyr::select(
    last_name, midl_name, name_sufx_cd, 
    sex, age,
    voter_status_desc, voter_status_reason_desc
    ) %>% 
  dplyr::arrange(voter_status_desc, voter_status_reason_desc, midl_name) %>% 
  knitr::kable()
```

-   Most are REMOVED, but some are ACTIVE and VERIFIED. That suggests
    the data entry for this record is done *after* verification.
-   Some appear to have the first name in the middle name field, e.g. (F
    M L) ("" "BRENDA" "PARRISH"), ("" "ALEXIS" "BULLARD")
-   Some appear to have first and middle names appended to the last
    name, e.g. (F M L) ("" "" "JONES LARRY MALLOR"), ("" ""
    "AMATO,KATHERINE,M")
-   Some are missing all the names!
-   Some appear to be test data, e.g. last name = XXX or "NEW TEST"

There are *very* few records missing first name or last name, and most
of them are REMOVED status. The easiest thing to do is just get rid of
those records.

**Exclude records with missing first or last name**

### Check for lower-case letters.

```{r}
d %>% dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "[a-z]"))

d %>% dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "[a-z]"))

d %>% dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "[a-z]"))
```

-   243 names with lower case letters.
-   Occur in last, first, and middle names.
-   Associated with particles where there would optionally be a space,
    e.g. De VANE.

**Map all letters to upper case**

### Digits

#### Check for digits

```{r}
d %>% dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "[0-9]"))

d %>% dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "[0-9]"))

d %>% dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "[0-9]"))
```

-   Zero substituted for O, e.g. J0HNSON, BURT0N
-   Some are obviously generation suffixes, e.g. ARGUS 4TH, LEAK 111
    (should be LEAK III)
-   Some are poor parsing into fields, e.g. MV 5/17/95 , MIZELLE25248249

Look at the digits individually.

#### Check for zero

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "0"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "0"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "0"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   270 names with zero
-   Occur in last, first, and middle names.
-   Most are zero substituted for O, e.g. J0HNSON, BURT0N
-   Some are pure numeric, e.g. 0, 01
-   Some are names with concatenated numeric, e.g. WAYNE030986,
    WRIGHT2106

**Map zero to O if name contains at least one letter and no digits 1-9**

#### Check for one

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "1"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "1"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "1"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   186 names with one
-   Occur in last, first, and middle names.
-   Most are 1 substituted for I in generation suffix, e.g. COX 1V, CARR
    111
-   Some are pure numeric, e.g. 01, 971
-   Some are wrongly parsed, e.g. MV 5/17/95
-   Some are names with concatenated numeric, e.g. LYNN2513, WRIGHT2106

**Delete generation suffixes where possible**

#### Check for two

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "2"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "2"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "2"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   15 names with two
-   Some are pure numeric, e.g. 205, 328
-   Some are names with concatenated numeric, e.g. LYNN1820, WRIGHT2106

#### Check for three

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "3"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "3"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "3"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   14 names with three
-   Some are pure numeric, e.g. 103, 328
-   Some are generation suffixes, e.g. 3RD., MACK 3RD
-   Some are names with concatenated numeric, e.g. LEE3708, SCOTT3450

#### Check for four

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "4"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "4"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "4"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   19 names with four
-   Some are pure numeric, e.g. 4625, 4932
-   Some are generation suffixes, e.g. ARGUS 4TH, MCREE 4
-   Some are names with concatenated numeric, e.g. DALE401, SCOTT3450
-   Some are intrusions in names, e.g. FR4ANK, MICHA4EL

#### Check for five

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "5"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "5"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "5"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   20 names with five
-   Some are pure numeric, e.g. 205, 2205
-   Some are generation suffixes, e.g. (NMN)5TH
-   Some are names with concatenated numeric, e.g. DALE401, SCOTT3450
-   Some are wrongly parsed, e.g. MV 5/17/95
-   Some are intrusions in names, e.g. FR4ANK, MICHA4EL
-   Some are substitution of 5 for S e.g. ALBER5TSON

#### Check for six

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "6"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "6"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "6"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   9 names with six
-   Some are pure numeric, e.g. 6, 4625
-   Some are names with concatenated nmeric, e.g. MICHAEL146, MICHAEL146

#### Check for seven

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "7"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "7"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "7"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   12 names with seven
-   Some are pure numeric, e.g. 491715, 971
-   Some are names with concatenated numeric, e.g. DALE401, SCOTT3450
-   Some are wrongly parsed, e.g. MV 5/17/95
-   Some are intrusions in names, e.g. JOYCE701, LOUIS7100

#### Check for eight

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "8"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "8"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "8"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   11 names with eight
-   Some are pure numeric, e.g. 328, 8017
-   Some are names with concatenated numeric, e.g. LEE3708, LYNN1820
-   Some are intrusions in names, e.g. J8IMMIE
-   Some might be substitution of 8 for SE e.g. BEA LOUI8

#### Check for nine

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "9"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "9"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "9"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   10 names with nine
-   Some are pure numeric, e.g. 971, 4932
-   Some are names with concatenated numeric, e.g. ANDERSON9104576,
    WAYNE030986
-   Some are wrongly parsed, e.g. MV 5/17/95
-   Some are intrusions in names, e.g. LO9UIS

#### Digit summary

**Map zero to O if there are any letters and no digits 1-9**

**One is sometimes substituted for "I" in generation suffixes. Remove
these suffixes from names.**

**Otherwise, map all digits to empty string.**

### Characters

#### Check for hyphens

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "-"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)
```

-   \~34k last names with hyphens
-   Look like legitimately hyphenated last names
-   Some hyphenated last names have an extra space, e.g. "TIPTON-
    BARNARD"

```{r}
x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "-"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)
```

-   \~5k first names with hyphens
-   Look like legitimately hyphenated first names

```{r}
x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "-"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   \~6k middle names with hyphens
-   Look like legitimately hyphenated middle names
-   Some hyphenated middle names have an extra space, e.g. "A - BERTHA",
    "H - JACK"

I suspect that hyphenation is likely to be a bit unreliable in
transcription.

**Map hyphen to empty string**

#### Check for slash

```{r}
d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "/"))

d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "/"))

d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "/"))

```

-   46 last names with slash
-   One obviously badly parsed - MV 5/17/95
-   Remainder being used equivalently to hyphen or indicating former
    name, e.g. MORRIS/BLOOM, WATERS/CRUZ

**Map slash to empty string**

#### Check for underscore

```{r}
d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(last_name, "_"))

d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(first_name, "_"))

d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(midl_name, "_"))
```

-   21 names with underscore
-   Being used equivalent to hyphen in one last name: SOLARZ_VOJDANI
-   Mostly used as equivalent of space in other names, e.g. KEVIN_C,
    DEAN\_\_ALAN

**Map underscore to empty string**

#### Check for percent

```{r}
d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(last_name, "%"))

d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(first_name, "%"))

d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(midl_name, "%"))
```

-   5 names with percent
-   Being used equivalent to hyphen in one last name: SCHERM%MARTIN
-   Possibly being used as substitute for E, e.g. DAV%, ANN%
-   Sometimes appears to be a pointless suffix, e.g. P%, EVE%

**Map percent to empty string**

#### Check for single quotes

```{r}
x <- d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "'"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::pull(last_name)

x <- d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "'"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::pull(first_name)

x <- d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "'"))
dim(x)
x %>%   
  dplyr::distinct() %>% 
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::pull(midl_name)
```

-   \~17k names with single quotes
-   Most look like correct names, e.g. O'NEAL, D'AGOSTINO
-   Some have a terminal quote, e.g. BONE', BOVA'
-   Some have extra space and/or a hyphen, e.g. "O' NEAL",
    "LACHARITE'-OTWELL"

I suspect that quotes are likely to be a bit unreliable in
transcription.

**Map single quote to empty string**

#### Check for double quotes

```{r}
d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "\""))

d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "\""))

d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "\""))
```

The backslashes are automatically inserted escaping so that the output
strings could be read as inputs without getting confused by the double
quotes.

-   24 names with double quotes
-   Some substitute double for single quote, e.g. the last name LA"BEE
    probably should have been LA'BEE
-   Some appear to be aliases or nicknames, e.g. MARY ("PETE"), "TATER"

**Map all double quotes to empty string**

#### Check for asterisk

```{r}
d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "\\*"))

d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "\\*"))

d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "\\*"))
```

-   31 names with asterisk
-   Asterisk substituted for single quote in last names, e.g. O*NEALS,
    D*AMICO
-   Asterisk used as a suffix in first and middle names

**Map asterisk to empty string**

#### Check for back-tick

```{r}
d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(last_name, "`"))

d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(first_name, "`"))

d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(midl_name, "`"))
```

-   114 names with back-tick
-   Some back ticks are substituted for single quotes, e.g.
    O`BRIANT, O`NEAL
-   Others are affixes for no obvious reason, e.g. CASEY`, ANNA`

**Map back-tick to empty string**

#### Check for tilde

```{r}
d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(last_name, "~"))

d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(first_name, "~"))

d %>% 
  dplyr::select(last_name, first_name, midl_name, sex) %>%
  dplyr::filter(stringr::str_detect(midl_name, "~"))
```

-   1 name with tilde
-   Being used equivalent to single quote, e.g. O\~CONNOR

**Map tilde to empty string**

#### Check for whitespace.

```{r}
x <- d %>% 
  dplyr::select(last_name, first_name, midl_name, sex, age) %>%
  dplyr::filter(stringr::str_detect(last_name, "\\s"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::select(last_name, first_name, midl_name, sex, age) %>%
  dplyr::filter(stringr::str_detect(first_name, "\\s"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::select(last_name, first_name, midl_name, sex, age) %>%
  dplyr::filter(stringr::str_detect(midl_name, "\\s"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()
```

-   111,836 names with whitespace
-   Some whitespace is because of prefixes, e.g. DA SILVA, LE BLANC
-   Some whitespace is separating generation suffixes, e.g. BENDER JR,
    HOLLERS 111
-   Some whitespace is instead of a hyphen, e.g. KROMIS BRESNIHAN, WATTS
    ST PIERREE
-   Some whitespace is incorrectly inserted, e.g MILLS- KHARBAT,
    REDFEARN- SHELTON
-   Some whitespace is probably variable between people, e.g. MC CADEN,
    MC COY
-   Some whitespace separates alias/nick/former names, e.g. PARISH
    (RAMON), SYKES (BRICKHOUSE)
-   Some whitespace is instead of a single quote, e.g. O BRIEN, O NEAL
-   Some whitespace is separating an honorific title, e.g. MRS CLAUD,
    MRS THEODORE
-   Some whitespace is separating a final honorific title, e.g. D MS

**Map whitespace to empty string**

#### Check for period

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "\\."))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "\\."))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "\\."))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()
```

-   10,017 names with period
-   Some are abbreviations of SAINT although spacing is inconsistent,
    e.g. ST. CLAIR, ST.CLAIR
-   Some are abbreviation of generation suffix, e.g. JR.
-   Some are substituted for whitespace, e.g. BRICE.MICHAEL ARTHUR
-   Some appear to be a random termination, e.g. PENDLETON-S. , SHIELDS.
-   Some indicate initials, e.g. W., J.C.
-   Some indicate a contraction, e.g. MARY ELIZ.
-   Some appear to substitute for a single quote, e.g. L.ORUNE OLDS

**Map period to empty string**

#### Check for comma

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, ","))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, ","))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, ","))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()
```

-   172 names with comma
-   Many commas are when multiple name components have been included in
    the last name, e.g. "AMATO,KATHERINE,M" "HODNETT,DORGIE,JR"
-   Some commas are when multiple name components have been included in
    the first name, e.g. "EUGENE, JR", ",JR"
-   Some commas are when multiple name components have been included in
    the first name, e.g. "MARY, D", "JAMES, II", "MAE, MRS", "W., SR.,"

**Map comma to empty string**

#### Check for backslash

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "\\\\"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "\\\\"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "\\\\"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()
```

-   74 names with backslash
-   Some appear to be terminal markers, e.g. BUFFKIN\\, KEVIN\\
-   Some are used as substitutes for whitespace, e.g. WILLIAM\\BOND,
    ROSE\\MERSON

**Map backslash to empty string**

#### Check for parentheses

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "[()]"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "[()]"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "[()]"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()
```

-   2,234 names with parentheses
-   Some appear to be maiden names, e.g. CARSON (WADE), MOCCIA (SMITH)
-   Some appear to be explanatory notes, e.g. NEESE (BLIND VOTER,
    JOHNSON (BLIND VOT
-   Some appear to indicate absence of a name, e.g. L(NN) = L (no name),
    (NMN) = (no middle name)
-   Some appear to be for honorific titles, e.g. FEE (SISTER), JOE (DR),
    (MRS)

**Map parentheses to empty string**

#### Check for braces

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "[{}]"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "[{}]"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()

x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "[{}]"))
dim(x)
x %>%   
  dplyr::slice_head(n = 100) %>% 
  dplyr::arrange(last_name, first_name, midl_name) %>% 
  knitr::kable()
```

-   5 names with braces
-   Some appear to be nicknames, e.g. MARY {HOLLY}, {ALEX}ANDER
-   Some might be OCR errors, e.g. D}\@IS = DENIS

**Map braces to empty string**

#### Check for other characters

```{r}
d %>% 
  dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "[^-a-zA-Z0-9/_%\'\"\\\\*\`~ \\.,\\\\(){}]"))

d %>% 
  dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "[^-a-zA-Z0-9/_%\'\"\\\\*\`~ \\.,\\\\(){}]"))

d %>% 
  dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "[^-a-zA-Z0-9/_%\'\"\\\\*\`~ \\.,\\\\(){}]"))
```

-   31 names with other characters
-   Some appear to substitute for single quote, e.g. O;NEAL, O=BOZOVICH
-   Some might be OCR errors, e.g. STAN;EY, E;OZABETH
-   Some are explanatory notes, e.g. "MERLE ATTN!!", "L!!!hold for",
    ""(DECEASED ??"
-   Some appear to be random junk, e.g. FRED\#, RORY], ";", "G\^"

**Map other characters to empty string**

#### Character summary

**Map non-alphanumeric characters to empty string**

### Words

Look for words that shouldn't be in names.

#### name_sufx_cd

`name_sufx_cd`: Voter name suffix

I am not going to use name suffix in entity resolution because age
should be sufficient and is much better quality.

Just look at what turns up in the name suffix in order to see what
occurs, so that the same values can be removed from the other name
fields where they shouldn't occur but do.

```{r}
d %>% dplyr::select(name_sufx_cd) %>% skimr::skim()
table(d$name_sufx_cd, useNA = "ifany")

# get a better look at the cleaned suffixes
d %>% 
  dplyr::mutate(
    sufx = name_sufx_cd %>% 
      stringr::str_to_upper() %>% 
      stringr::str_remove_all(pattern = "[^A-Z0-9]") %>% # remove non-alphanumeric
      dplyr::na_if("") 
  ) %>% 
  dplyr::count(sufx) %>% 
  dplyr::arrange(desc(n), sufx) %>% 
  knitr::kable()
```

-   There are honorific titles: MRS, MR, MS, REV, MD, DR, JD
-   There are generation suffixes: JR, SR, I (1, J, L), II (2, 2ND, 11,
    LL), III (3RD, 111, IIL, ILI, LLL), IV, V (5TH), VI, VII, 8TH

#### Honorific in name

Look for honorifics that have been put in name fields.

##### Last name honorific

```{r}
# last name
words <- c(
  "MR", "MISTER", "MASTER", "MRS", "MS", "MISS", 
  "REV", "REVEREND", "SR", "SISTER", "BR", "BROTHER",
  "FATHER", "MOTHER", "PASTOR", "ELDER", "BISHOP",
  "DR", "DOCTOR", "MD", "JD", "PROF", "PROFESSOR"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    last_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(last_name, sex, first_name)
nrow(x)
x %>% knitr::kable()
```

Last name

-   BROTHER, CAPTAIN, COLONEL, DOCTOR, MASTER, MISS, MISTER appear to be
    legitimate last names
-   MR is the only prefix: MR FEWEL
-   SR, MD, SISTER, DR are suffixes

##### First name honorific

```{r}
# first name
words <- c(
  "MR", "MISTER", "MASTER", "MRS", "MS", "MISS", 
  "REV", "REVEREND", "SR", "SISTER", "BR", "BROTHER",
  "FATHER", "MOTHER", "PASTOR", "ELDER", "BISHOP",
  "DR", "DOCTOR", "MD", "JD", "PROF", "PROFESSOR"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    first_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(first_name, sex, last_name)
nrow(x)
x %>% knitr::kable()
```

First name

-   Apparently legitimate first names: BROTHER, DOCTOR, MASTER
-   As entire first name: MRS, REV, DR, MISS, MISTER, PROFESSOR, SISTER,
    SR
-   Prefixes: MRS, REV, DR, SISTER
-   Suffixes: MRS, SR

##### Middle name honorific

```{r}
# middle name
words <- c(
  "MR", "MISTER", "MASTER", "MRS", "MS", "MISS", 
  "REV", "REVEREND", "SR", "SISTER", "BR", "BROTHER",
  "FATHER", "MOTHER", "PASTOR", "ELDER", "BISHOP",
  "DR", "DOCTOR", "MD", "JD", "PROF", "PROFESSOR"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    midl_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(midl_name, sex, last_name)
nrow(x)
x %>% knitr::kable()
```

Middle name

-   Apparently legitimate middle names: MASTER
-   As entire middle name: MRS, BROTHER, DOCTOR, DR, MD, MISS, MISTER,
    MS, PROFESSOR, REV, REVEREND, SISTER, SR
-   Prefixes: MRS
-   Suffixes: MRS, SR, SISTER, REV, DR, MS, MR, MISS, MD, BR, MASTER

#### Generation suffix in name

Look for generation suffixes that have been put in name fields.

##### Last name generation

```{r}
# last name
words <- c(
  "JR", "JNR", "JUNIOR", "SR", "SNR", "SENIOR",
  "1ST", "2ND", "3RD", "4TH", "5TH", "6TH", "7TH", "8TH",
  "FIRST", "SECOND", "THIRD", "FOURTH", "FIFTH", "SIXTH", "SEVENTH", "EIGHTH", "EIGHTTH",
  "1", "2", "3", "4", "5", "6", "7", "8",
  "[I1JL]{1,3}", "[I1JL]V", "V[I1JL]{1,3}"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    last_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(last_name, sex, first_name)
nrow(x)
x %>% knitr::kable()
```

Last name

-   Appear to be legitimate last names:
-   As entire last name:
-   Prefixes:
-   Suffixes:

##### First name generation

```{r}
# first name
words <- c(
  "JR", "JNR", "JUNIOR", "SR", "SNR", "SENIOR",
  "1ST", "2ND", "3RD", "4TH", "5TH", "6TH", "7TH", "8TH",
  "FIRST", "SECOND", "THIRD", "FOURTH", "FIFTH", "SIXTH", "SEVENTH", "EIGHTH", "EIGHTTH",
  "1", "2", "3", "4", "5", "6", "7", "8",
  "[I1JL]{1,3}", "[I1JL]V", "V[I1JL]{1,3}"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    first_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(first_name, sex, last_name)
nrow(x)
x %>% knitr::kable()
```

First name

-   Appear to be legitimate first names:
-   As entire first name:
-   Prefixes:
-   Suffixes:
-   Suffixes: MRS, SR

##### Middle name generation

```{r}
# middle name
words <- c(
  "JR", "JNR", "JUNIOR", "SR", "SNR", "SENIOR",
  "1ST", "2ND", "3RD", "4TH", "5TH", "6TH", "7TH", "8TH",
  "FIRST", "SECOND", "THIRD", "FOURTH", "FIFTH", "SIXTH", "SEVENTH", "EIGHTH", "EIGHTTH",
  "1", "2", "3", "4", "5", "6", "7", "8",
  "[I1JL]{1,3}", "[I1JL]V", "V[I1JL]{1,3}"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    midl_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(midl_name, sex, last_name)
nrow(x)
x %>% knitr::kable()
```

Middle name

-   Appear to be legitimate middle names:
-   As entire middle name:
-   Prefixes:
-   Suffixes:

#### No name

```{r}
# last name
words <- c(
  "NN", "NMN", "NAME"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    last_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(last_name, sex, first_name)
nrow(x)
x %>% knitr::kable()
```

Last name

-   Apparently legitimate last name: "VAN NAME"

```{r}
# first name
words <- c(
  "NN", "NMN", "NAME"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    first_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(first_name, sex, last_name)
nrow(x)
x %>% knitr::kable()
```

First name

-   As entire first name: "NO OTHER NAME"
-   Suffixes: NN, NMN

```{r}
# middle name
words <- c(
  "NN", "NMN", "NAME"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    midl_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(midl_name, sex, last_name)
nrow(x)
x %>% knitr::kable()
```

Middle name

-   As entire middle name: NMN, NN
-   Prefixes: NMN
-   Suffixes: NN

#### Unknown etc.

```{r}
# last name
words <- c(
  "UNK", "UNKNOWN", "AKA", "KNOWN AS", "ALSO KNOWN AS", "ALIAS",
  "BLIND"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    last_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(last_name, sex, first_name)
nrow(x)
x %>% knitr::kable()
```

Last name

-   Apparently legitimate last name: UDO-AKA
-   As entire last name: UNKNOWN

```{r}
# first name
words <- c(
  "UNK", "UNKNOWN", "AKA", "KNOWN AS", "ALSO KNOWN AS", "ALIAS",
  "BLIND"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    first_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(first_name, sex, last_name)
nrow(x)
x %>% knitr::kable()
```

First name

-   As entire first name: UNKNOWN

```{r}
# middle name
words <- c(
  "UNK", "UNKNOWN", "AKA", "KNOWN AS", "ALSO KNOWN AS", "ALIAS",
  "BLIND"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") %>% 
  glue::glue(x = . , "({x})")

x <- d %>% 
  dplyr::filter(
    midl_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(midl_name, sex, last_name)
nrow(x)
x %>% knitr::kable()
```

Middle name

-   Apparently legitimate middle name: AKA (although that's implausible)

#### Test cases.

Find test records

```{r}
# last name
words <- c(
  "TEST", "TST", "DUMMY", "VOTER",  "([A-Z])\\1{2,}"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|")

x <- d %>% 
  dplyr::select(last_name, first_name, midl_name, name_sufx_cd, sex, age, res_city_desc, street_name, house_num, area_cd, phone_num) %>% 
  dplyr::filter(
    last_name %>% 
      stringr::str_to_upper() %>%
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>%
      stringr::str_squish() %>%
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(last_name, res_city_desc, street_name, sex, first_name)
nrow(x)
x %>% knitr::kable()
```

Last name

-   Apparently legitimate middle names:
-   As entire middle name:
-   Prefixes:
-   Suffixes:

```{r}
# first name
words <- c(
  "THIS[A-Z]*", "TEST", "TST", "DUMMY", "FAKE", "FALSE", "BOGUS", "VOTER",  "([A-Z])\\1{2,}"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|")

x <- d %>% 
  dplyr::select(last_name, first_name, midl_name, name_sufx_cd, sex, age, res_city_desc, street_name, house_num, area_cd, phone_num) %>% 
  dplyr::filter(
    first_name %>% 
      stringr::str_to_upper() %>%
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>%
      stringr::str_squish() %>%
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(first_name, last_name, res_city_desc, street_name, sex)
nrow(x)
x %>% knitr::kable()
```

First name

-   Apparently legitimate middle names:
-   As entire middle name:
-   Prefixes:
-   Suffixes:

```{r}
# middle name
words <- c(
  "THIS[A-Z]*", "TEST", "TST", "DUMMY", "FAKE", "FALSE", "BOGUS", "VOTER",  "([A-Z])\\1{2,}"
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|")

x <- d %>% 
  dplyr::select(last_name, first_name, midl_name, name_sufx_cd, sex, age, res_city_desc, street_name, house_num, area_cd, phone_num) %>% 
  dplyr::filter(
    midl_name %>% 
      stringr::str_to_upper() %>%
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>%
      stringr::str_squish() %>%
      stringr::str_detect(pattern = words)
  ) %>% 
  dplyr::arrange(midl_name, last_name, res_city_desc, street_name, sex, first_name)
nrow(x)
x %>% knitr::kable()
```

Middle name

-   Apparently legitimate middle names:
-   As entire middle name:
-   Prefixes:
-   Suffixes:


# Clean name variables

The aggregated cleaning suggestions are:

+-------------+-------------+-------------+-------------+-------------+
| Issue       | `last_name` | `           | `midl_name` | Action      |
|             |             | first_name` |             |             |
+=============+:===========:+:===========:+:===========:+=============+
| Missing     | 122         | 254         | 553,015     | Exclude     |
|             |             |             |             | record if   |
|             |             |             |             | first or    |
|             |             |             |             | last name   |
|             |             |             |             | missing     |
+-------------+-------------+-------------+-------------+-------------+
| Lower case  | 50          | 24          | 169         | Map all     |
| letters     |             |             |             | letters to  |
|             |             |             |             | upper case  |
+-------------+-------------+-------------+-------------+-------------+
| Digits      | 90          | 81          | 299         | Map digits  |
|             |             |             |             | to empty    |
|             |             |             |             | string if   |
|             |             |             |             | not         |
|             |             |             |             | otherwise   |
|             |             |             |             | mapped      |
+-------------+-------------+-------------+-------------+-------------+
| Zero        | 67          | 73          | 130         | Map zero to |
|             |             |             |             | O if name   |
|             |             |             |             | contains at |
|             |             |             |             | least one   |
|             |             |             |             | letter and  |
|             |             |             |             | no digits   |
|             |             |             |             | 1-9         |
+-------------+-------------+-------------+-------------+-------------+
| One         | 20          | 3           | 163         |             |
+-------------+-------------+-------------+-------------+-------------+
| Two         | 1           | 1           | 13          |             |
+-------------+-------------+-------------+-------------+-------------+
| Three       | 1           | 0           | 13          |             |
+-------------+-------------+-------------+-------------+-------------+
| Four        | 3           | 1           | 15          |             |
+-------------+-------------+-------------+-------------+-------------+
| Five        | 3           | 0           | 17          |             |
+-------------+-------------+-------------+-------------+-------------+
| Six         | 1           | 1           | 7           |             |
+-------------+-------------+-------------+-------------+-------------+
| Seven       | 4           | 0           | 8           |             |
+-------------+-------------+-------------+-------------+-------------+
| Eight       | 0           | 2           | 9           |             |
+-------------+-------------+-------------+-------------+-------------+
| Nine        | 4           | 0           | 6           |             |
+-------------+-------------+-------------+-------------+-------------+
| Hyphen      | 34,325      | 5,298       | 6,304       | Map hyphen  |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Slash       | 46          | 9           | 1,032       | Map slash   |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Single      | 9,712       | 1,965       | 5,426       | Map single  |
| quote       |             |             |             | quote to    |
|             |             |             |             | empty       |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Double      | 1           | 4           | 19          | Map double  |
| quote       |             |             |             | quote to    |
|             |             |             |             | empty       |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Asterisk    | 7           | 1           | 23          | Map         |
|             |             |             |             | asterisk to |
|             |             |             |             | empty       |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Back tick   | 10          | 71          | 33          | Map back    |
|             |             |             |             | tick to     |
|             |             |             |             | empty       |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Tilde       | 1           | 0           | 0           | Map tilde   |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Underscore  | 1           | 17          | 3           | Map         |
|             |             |             |             | underscore  |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Percent     | 1           | 4           | 0           | Map percent |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Whitespace  | 13,637      | 23,789      | 74,410      | Map         |
|             |             |             |             | whitespace  |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Period      | 44          | 651         | 9,322       | Map period  |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Comma       | 63          | 51          | 58          | Map comma   |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Backslash   | 4           | 3           | 67          | Map         |
|             |             |             |             | backslash   |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Parentheses | 22          | 105         | 2,107       | Map         |
|             |             |             |             | parentheses |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Braces      | 0           | 1           | 4           | Map braces  |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+
| Other       | 8           | 8           | 15          | Map other   |
| characters  |             |             |             | characters  |
|             |             |             |             | to empty    |
|             |             |             |             | string      |
+-------------+-------------+-------------+-------------+-------------+

: Name cleaning suggestions

upcase\
map non-alphanumeric to space\
remove nn, nmn, etc.\
remove terminal generation suffixes\
remove honorific prefixes and suffixes\
map all space to null

```{r}
knitr::knit_exit()
```

## Upper case

Map all letters to upper case.

```{r}
d <- d %>% 
  dplyr::mutate_if(is.character, stringr::str_to_upper)
```

**Move JR suffix from `last_name` to `name_sufx_cd`**

\*\*Map terminal 1\* in `last_name` to I\*

## Double quotes

Map double quotes to single quotes.

```{r}
d <- d %>% 
  dplyr::mutate_if(is.character, function(x) stringr::str_replace_all(x, "\"", "'"))
```

## Name substitutions

These single character substitutions are only applied in names.

0 ==\> O\
\* ==\> '\
. ==\> space\
, ==\> space\
[1-9] ==\> null ==\> null

```{r}
d <- d %>% 
  dplyr::mutate_at(
    vars(ends_with("_name")), 
    function(x) stringr::str_replace_all(x, c(
      "0" = "O",
      "\\*" = "'",
      "\\." = " ",
      "," = " ",
      "[1-9]" = "",
      "\\\\"= ""
    ))
  )
```

## Excess whitespace

Replace consecutive whitespace with a single space and remove leading
and trailing spaces.

```{r}
d <- d %>% 
  dplyr::mutate_if(is.character, stringr::str_squish())
```

# Characterise data (active verified records)

**From this point on, limit observations to: `voter_status_desc` ==
"ACTIVE" & `voter_status_reason_desc` == "VERIFIED"**

Only keep columns which might conceivably be of some use.

```{r}
d <- d %>% 
  dplyr::filter(voter_status_desc == "ACTIVE" & voter_status_reason_desc == "VERIFIED") %>% 
  dplyr::select(
    county_id, voter_reg_num,
    last_name : street_sufx_cd, unit_designator : zip_code,
    area_cd, phone_num,
    sex : registr_dt
  )
```

```{r}
glimpse(d)
skimr::skim(d)
```

-   There are no warning messages from `skim()`, so the records with
    unusual characters have been excluded.
-   \~4.1M observations

## county_id

`county_id`: County identification number

```{r}
d %>% dplyr::select(county_id) %>% skimr::skim()
```

-   Never missing
-   Integer 1 .. 100

## voter_reg_num

`voter_reg_num`: Voter registration number (unique by county)

```{r}
d %>% dplyr::select(voter_reg_num) %>% skimr::skim()
d %>% dplyr::select(voter_reg_num) %>% dplyr::mutate(voter_reg_num = as.integer(voter_reg_num)) %>% skimr::skim()
```

-   \~1.8M unique values
-   Never missing
-   Integer 1 .. \~400M

## id

`id`: Unique person identifier

Add person `id`. Assume that each record corresponds to a unique person
because the source is a point-in-time snapshot, the metadata claims
`voter_reg_num` is unique within `county_id`, electoral officials
presumably make serious efforts to find and remove duplicates, and we
have restricted to only active verified records. (In practice, I
wouldn't be surprised if there were a small number of duplicates.)

```{r}
d <- d %>% 
  dplyr::select(-voter_reg_num) %>% # drop voter_reg_num; no longer needed
  dplyr::mutate(id = sample.int(nrow(d))) %>%  # id is a random permutation of row numbers
  dplyr::arrange(id) # randomise row order in case it's systematic

d %>% dplyr::select(id) %>% skimr::skim()
```

-   \~4.1M unique values
-   Integer 1 .. \~4.1M
-   No duplicate values
