---
title: "03-1_name_freq_eql_global"
subtitle: "Name frequency induced by equality - Global analysis"
author: "Ross Gayler"
date: "2021-01-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup}
# Set up the project environment, because each Rmd file knits in a new R session
# so doesn't get the project setup from .Rprofile

# Project setup
library(here)
source(here::here("code", "setup_project.R"))

# Extra set up for the 02*.Rmd notebooks
# source(here::here("code", "setup_03.R"))

# Extra set up for this notebook
library(zipfR)

# start the execution time clock
tictoc::tic("Computation time (excl. render)")
```

# Introduction

This notebook (`02-2_mk_block_vars`) constructs the potential blocking
variables identified in the previous notebook:

-   (sex, age)
-   (age, county)
-   (sex, age, county)


# Read data

Read the usable data. Remember that this consists of only the ACTIVE &
VERIFIED records.

```{r}
# Show the entity data file location
# This is set in code/file_paths.R
fs::path_file(f_entity_cln_fst)

# get entity data
d <- fst::read_fst(
  f_entity_cln_fst#,
  # columns = c("id", "sex", "age_cln", "age_cln_miss", "birth_place", "county_id")
  ) %>% 
  tibble::as_tibble()

dim(d)
```

# xxx

Identify the missing values in the blocking variables and flag those
records as exclusions for blocking and modelling.

Sex and age have missing values, county ID does not.

```{r}

tfl_last_name <- zipfR::vec2tfl(d$last_name_cln)

tfl_last_name

plot(tfl_last_name, log = "xy")

spc_last_name <- tfl2spc(tfl_last_name)

spc_last_name

summary(spc_last_name)

plot(spc_last_name, log = "xy")

vgc_last_name <- zipfR::vec2vgc(d$last_name_cln, m.max = 9)

vgc_last_name

summary(vgc_last_name)

plot(vgc_last_name, add.m = 1:9)

abline(0, 1)



```

-   (age, county) excludes \~31k (\~0.8%) of records from blocking and
    modelling.
-   (sex, age) and (sex, age, county) excludes \~47k (\~1.1%) of records
    from blocking and modelling.

# Save enity blocking variables

Save the blocking variables as a separate file that can be joined to the clean entity data.

```{r}
# # Show the clean data file location
# # This is set in code/file_paths.R
# fs::path_file(f_entity_blk_fst)
# 
# # save the entity blocking variables (cheap-skate caching)
# d %>% 
#   dplyr::select(id, starts_with(c("excl_blk_", "blk_"))) %>% 
#   fst::write_fst(f_entity_blk_fst, compress = 100) %>% 
#   dplyr::glimpse()

```

# Timing {.unnumbered}

```{r echo=FALSE}
tictoc::toc()
```
