---
title: "03-1_name_freq_eql_global"
subtitle: "Name frequency induced by equality - Global analysis"
author: "Ross Gayler"
date: "2021-01-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup}
# Set up the project environment, because each Rmd file knits in a new R session
# so doesn't get the project setup from .Rprofile

# Project setup
library(here)
source(here::here("code", "setup_project.R"))

# Extra set up for the 03*.Rmd notebooks
# source(here::here("code", "setup_03.R"))

# Extra set up for this notebook
library(zipfR)

# start the execution time clock
tictoc::tic("Computation time (excl. render)")
```

# Introduction

This series of notebooks (`03`) only considers the equality relation on names.
That is, two names are either identical or equal.
This can be considered as a very coarse and extreme quantisation of an underlying continuous similarity relation.

This notebook (`03-1_name_freq_eql_global`) looks at the frequencies of names in the entire identity data set.

The basic idea of this project is that the information provided by a given level of name similarity between a query record and a dictionary record depends on the frequency of the names.
High similarity (equality) between the query and dictionary records is presumed to be less informativeif the names are common in the data.

# Read entity data

Read the usable entity data. Remember that this consists of only the ACTIVE &
VERIFIED records.

The analyses in this notebook do not involve blocking,
so ignore the missing value indicators for the blocking records.
The name frequencies will be based on all usable records
regardless of whether those records have missing blocking values.

```{r}
# Show the entity data file location
# This is set in code/file_paths.R
fs::path_file(f_entity_cln_fst)

# get entity data
d <- fst::read_fst(
  f_entity_cln_fst#,
  # columns = c("id", "sex", "age_cln", "age_cln_miss", "birth_place", "county_id")
  ) %>% 
  tibble::as_tibble()

dim(d)
```

# xxx

Identify the missing values in the blocking variables and flag those
records as exclusions for blocking and modelling.

Sex and age have missing values, county ID does not.

```{r}

tfl_last_name <- zipfR::vec2tfl(d$last_name_cln)

tfl_last_name

plot(tfl_last_name, log = "xy")

spc_last_name <- tfl2spc(tfl_last_name)

spc_last_name

summary(spc_last_name)

plot(spc_last_name, log = "xy")

vgc_last_name <- zipfR::vec2vgc(d$last_name_cln, m.max = 9)

vgc_last_name

summary(vgc_last_name)

plot(vgc_last_name, add.m = 1:9)

abline(0, 1)



```

-   (age, county) excludes \~31k (\~0.8%) of records from blocking and
    modelling.
-   (sex, age) and (sex, age, county) excludes \~47k (\~1.1%) of records
    from blocking and modelling.

# Save enity blocking variables

Save the blocking variables as a separate file that can be joined to the clean entity data.

```{r}
# # Show the clean data file location
# # This is set in code/file_paths.R
# fs::path_file(f_entity_blk_fst)
# 
# # save the entity blocking variables (cheap-skate caching)
# d %>% 
#   dplyr::select(id, starts_with(c("excl_blk_", "blk_"))) %>% 
#   fst::write_fst(f_entity_blk_fst, compress = 100) %>% 
#   dplyr::glimpse()

```

# Timing {.unnumbered}

```{r echo=FALSE}
tictoc::toc()
```
