---
title: "[meta] Project Workflow Status"
subtitle: "m_00_status"
author: "Ross Gayler"
date: "`r date()`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
<<<<<<< HEAD
=======
  markdown: 
    wrap: 72
>>>>>>> 0e893b2b6ac0165542d0bbdae0e09af26c155cf8
---

```{r, echo = FALSE}
library(targets)
```

This notebook displays the status of the project workflow, that is,
whether everything is computationally up to date. It displays the status
as at the time the notebook is executed, so if the status has changed
since this notebook was last executed that fact won't be reflected in
this rendered notebook.

This notebook is currently intended to be run manually when you feel the
need, for example, when returning to the project.

# `{renv}` status

`{renv}` tracks the used `R` packages and their versions.

```{r renv}
# Get {targets} to tell {renv} which packages it uses
targets::tar_renv() 

renv::status()
```

# `{targets}` status

`{targets}` is used to track the computational dependencies between the
project components and orchestrate recalculation when any components are
changed.

-   `c_*` components are in the core pipeline
-   `p_*` components are in the publications pipelines
-   `m_*` components are in the meta pipelines

Functions may be used across multiple pipeline categories, so function
names are not prefixed.

Meta pipelines generally originate from a target on the core pipeline.
Each meta pipeline generally explores some aspect of that source target
which informs the design of the next stage of processing on the core
pipeline.

Each of the meta pipelines has leaves that are reports rendered from
`Rmd` files that are managed jointly by `{targets}` and `{workflowr}`.
`{targets}` is aware of the upstream dependencies (data and functions)
and the `Rmd` file itself. If any of these is invalidated `tar_make()`
will rebuild the report using `wflow_build()`.

`{targets}` is *not* aware of the rendered meta report files. The
rendered report files are managed exclusively by `{workflowr}`. This
means that you can use `{workflowr}` to build and publish the report
without making `{targets}` think that the report is out of date because
`{workflowr}` has re-rendered it.

```{r targets}
# outline
targets::tar_glimpse()

# core detail
targets::tar_visnetwork(allow = starts_with("c_"))

# meta detail
targets::tar_visnetwork(allow = starts_with("m_"))

# excruciating detail
targets::tar_visnetwork()
```

# `{workflowr}` status

`{workflowr}` is used to render the *meta* documents
(`analysis/m_*.Rmd`) and build the project website.

`{workflowr}` is only concerned with the modification and `git` commit
dates of the `Rmd` and `HTML` files, whereas `{targets}` is concerned
with the computational dependencies. Thus, you should rely on the
`{targets}` status to ensure the meta reports are computationally up to
date, then rely on the `{workflowr}` status to manage the building and
publication of the entire website.

```{r workflowr}
workflowr::wflow_status()
```
