---
title: "[meta] Check name variables"
subtitle: "m_01_8_check_name"
author: "Ross Gayler"
date: "2021-05-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup}
# NOTE this notebook can be run manually or automatically by {targets}
# So load the packages required by this notebook here
# rather than relying on _targets.R to load them.

# Set up the project environment, because {workflowr} knits each Rmd file 
# in a new R session, and doesn't execute the project .Rprofile

library(targets) # access data from the targets cache

library(tictoc) # capture execution time
library(here) # construct file paths relative to project root
library(fs) # file system operations
library(dplyr) # data wrangling
library(gt) # table formatting
library(stringr) # string matching
library(vroom) # fast reading of delimited text files
library(lubridate) # date parsing
library(forcats) # manipulation of factors
library(ggplot2) # graphics
library(skimr) # compact summary of each variable
library(tidyr) # data tidying
library(glue) # string interpolation

# start the execution time clock
tictoc::tic("Computation time (excl. render)")

# Get the path to the raw entity data file
# This is a target managed by {targets}
f_entity_raw_tsv <- tar_read(c_raw_entity_data_file)
```

```{r, include = FALSE, cache = FALSE}
# Get all the functions used by {targets} 
# so we can display and use them in this notebook
# with a guarantee that they will be identical
# here and when used by {targets}.

# This is  work-around because tarchetypes::tar_knitr_deps()
# fails if a call to here::here() is embedded in 
# a call to knitr::read_chunk()
knitr::read_chunk("R/functions.R")
```

# Introduction

The aim of this set of meta notebooks is to work out how to read the raw
entity data. and get it sufficiently neatened so that we can construct
standardised names and modelling features without needing any further
neatening. To be clear, the target (`c_raw_entity_data`) corresponding
to the objective of this set of notebooks is the neatened raw data,
before constructing any modelling features.

This notebook documents the checking of the person name variables for
any issues that need fixing.

These variables will be used to construct the main predictors in the
compatibility models.

Regardless of whether there are any issues that need to be fixed, the
analyses here may inform our use of these variables in later analyses.

Define the name variables.

-   `last_name` - Voter last name
-   `first_name` - Voter first name
-   `midl_name` - Voter middle name
-   `name_sufx_cd` - Voter name suffix

```{r}
vars_name <- c(
  "last_name", "first_name", "midl_name", "name_sufx_cd" 
)
```

# Read entity data

Read the raw entity data file using the previously defined functions
`raw_entity_data_read()`, `raw_entity_data_excl_status()`,
`raw_entity_data_excl_test()`, `raw_entity_data_drop_novar()`,
`raw_entity_data_parse_dates()`, `raw_entity_data_drop_admin()`,
and `raw_entity_data_drop_demog()`.

```{r, raw_entity_data_read, echo = FALSE}
```

```{r, raw_entity_data_excl_status, echo = FALSE}
```

```{r, raw_entity_data_excl_test, echo = FALSE}
```

```{r, raw_entity_data_drop_novar, echo = FALSE}
```

```{r, raw_entity_data_parse_dates, echo = FALSE}
```

```{r, raw_entity_data_drop_admin, echo = FALSE}
```

```{r, raw_entity_data_drop_demog, echo = FALSE}
```

```{r}
# Show the data file name
fs::path_file(f_entity_raw_tsv)

d <- raw_entity_data_read(f_entity_raw_tsv) %>% 
  raw_entity_data_excl_status() %>% 
  raw_entity_data_excl_test() %>% 
  raw_entity_data_drop_novar() %>% 
  raw_entity_data_parse_dates() %>% 
  raw_entity_data_drop_admin() %>% 
  raw_entity_data_drop_demog()

dim(d)
```

Take a quick look at the distributions.

```{r}
d %>% 
  dplyr::select(all_of(vars_name)) %>% 
  skimr::skim()
```

-   `last_name` 100% filled
-   `first_name` \~100% filled (23 missing)
-   `midl_name` 94% filled
-   `name_sufx_cd` 6% filled

# Name length

Look at the distributions of name lengths first, before moving on to
analyses more focused on standardisation.

Calculate the lengths of the name variables.

```{r}
d <- d %>% 
  dplyr::mutate(
    len_last = stringr::str_length(last_name),
    len_first = stringr::str_length(first_name),
    len_midl = stringr::str_length(midl_name)
  )
```

## last_name

-   `last_name` - Voter last name

Look at the distributions of name lengths.

```{r}
summary(d$len_last)

d %>% 
  dplyr::count(len_last) %>% 
  dplyr::arrange(len_last) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)

d %>% 
  ggplot() +
  geom_histogram(aes(x = len_last), binwidth = 1) +
  scale_y_sqrt()
```

Look at examples of short names.

```{r}
# length == 1
d %>% 
  dplyr::filter(len_last == 1) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   1-letter last names are very rare
-   Most 1-letter last names are probably errors

```{r}
# length == 2
d %>% 
  dplyr::filter(len_last == 2) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>%
  dplyr::arrange(last_name, first_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   Most 2-letter last names are probably valid.

Look at examples of long names.

```{r}
# length == 21
d %>% 
  dplyr::filter(len_last == 21) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   21-letter last names are hyphenated

```{r}
# length == 20
d %>% 
  dplyr::filter(len_last == 20) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>%
  dplyr::arrange(last_name, first_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   20+-letter last names appear to be multi-word and/or hyphenated

## first_name

-   `first_name` - Voter first name

Look at the distributions of name lengths.

```{r}
summary(d$len_first)

d %>% 
  dplyr::count(len_first) %>% 
  dplyr::arrange(len_first) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)

d %>% 
  ggplot() +
  geom_histogram(aes(x = len_first), binwidth = 1) +
  scale_y_sqrt()
```

-   Possibly more 1-letter first names than I would expect

Look at the missing names.

```{r}
d %>% 
  dplyr::filter(is.na(first_name)) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::arrange(last_name, first_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   Some missing first names look like the middle name is actually the
    first name, e.g. ? JASON ALEXANDER
-   Some missing first names appear to have only a last name, e.g. ? ?
    AMEN, ? ? GRAYWOLF, ? ? SILVERMOON
-   Some missing first names appear to have the entire name in the last
    name variable, e.g. ? ? FRYE WILLIAM C

Look at examples of short names.

```{r}
# length == 1
d %>% 
  dplyr::filter(len_first == 1) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::slice_sample(n = 20) %>%
  dplyr::arrange(last_name, first_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   The 1-letter first names appear to be using an initial as the first
    name

```{r}
# length == 2
d %>% 
  dplyr::filter(len_first == 2) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 50) %>%
  dplyr::arrange(first_name, last_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

2-letter first names appear to be:

-   Valid, e.g. JO ANNE COREY, ED ? HODGE
-   An initial with punctuation, e.g. A. L LITTLE, D. A. WHITNER
-   Multiple initials run together, e.g. GW ? BIVINS, JP ? DENTON
-   Part of a multi word name that has bee split across the first and
    middle name variables, e.g. LA SONDA FOWLER

Look at the long names.

```{r}
# length >= 16
d %>% 
  dplyr::filter(len_first >= 16) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 50) %>%
  dplyr::arrange(first_name, last_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

Long first names appear to be:

-   Long non-anglo names, e.g. LAKSHMINARAYANAN
-   Multi-word and/or hyphenated, e.g. ELIZABETH-LINDSAY

## midl_name

-   `midl_name` - Voter middle name

These names will often be missing or initials only.

Look at the distributions of name lengths.

```{r}
summary(d$len_midl)

d %>% 
  dplyr::count(len_midl) %>% 
  dplyr::arrange(len_midl) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)

d %>% 
  ggplot() +
  geom_histogram(aes(x = len_midl), binwidth = 1) +
  scale_y_sqrt()
```

-   *Many* records are missing middle name
-   Spike of 1-letter names will be initials

Look at the long names.

```{r}
# length >= 16
d %>% 
  dplyr::filter(len_midl >= 16) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 50) %>%
  dplyr::arrange(midl_name, last_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   Long middle names appear to be multiple names and/or hyphenated
-   Some long middle names are parenthesised expansions of first and
    middle names that were given as initials, e.g. A "E (ALONZO EDWARD)"
    LITTLE

# Name suffix code

-   `name_sufx_cd` - Voter name suffix

This is intended for generation markers, e.g. Junior, Senior.

I am not going to use name suffix in entity resolution because age
should be sufficient and is much better quality. I will look at what
values turn up in the name suffix because the same values sometimes
wrongly occur in the main name variables. Knowing what values occur may
help us to remove those values from the main name variables.

```{r}
d %>% dplyr::select(name_sufx_cd) %>% skimr::skim()

table(d$name_sufx_cd, useNA = "ifany") %>% sort() %>% rev()

d %>% 
  dplyr::count(name_sufx_cd) %>% 
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)


# get a better look at the cleaned suffixes
d %>% 
  dplyr::mutate(
    sufx = name_sufx_cd %>% 
      stringr::str_to_upper() %>% 
      stringr::str_remove_all(pattern = "[^A-Z0-9]") %>% # remove non-alphanumeric
      dplyr::na_if("") 
  ) %>% 
  dplyr::count(sufx) %>% 
  dplyr::filter(n > 1) %>% 
  dplyr::arrange(desc(n), sufx) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   Only \~6% of records have a name suffix code
-   There are generation suffixes: JR, SR, I, II (11), III (111), IV, V,
    VI, VII
-   There are honorific titles: MRS, MR, MS, DR, REV
-   The field lentgh appears to be hard-limited to three characters,
    e.g. (JR, (II

# Standardisation

Look at issues that might be addressed by standardisation.

For each type of standardisation issue look at first middle and last
names separately, because the issue may manifest differently in each of
the name variables.

## Lower-case letters

The alphabetic values are almost entirely upper case. Check for lower
case letters.

```{r}
d %>% dplyr::select(last_name) %>%
  dplyr::filter(stringr::str_detect(last_name, "[a-z]"))

d %>% dplyr::select(first_name) %>%
  dplyr::filter(stringr::str_detect(first_name, "[a-z]"))

d %>% dplyr::select(midl_name) %>%
  dplyr::filter(stringr::str_detect(midl_name, "[a-z]"))
```

-   Lower case letters occur in last, first, and middle names
-   Associated with particles where there would optionally be a space,
    e.g. JoANN, McBride

## Non-alphanumeric

Check for non-alphanumeric characters in names.

### Hyphen

Check for hyphens.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "-"))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~21k last names with hyphens
-   Look like valid hyphenated last names

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "-"))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~3k first names with hyphens
-   Look like valid hyphenated first names

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "-"))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~4k middle names with hyphens
-   Look like valid hyphenated middle names

### Single quote

Check for single quotes.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "'"))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~5k last names with single quotes
-   Most look like legitimately quoted last names
-   Some look like typos, e.g. LEA'KES, PERRY'KNIGHT
-   Some look like they were intended to be diacritics, e.g. GORRE'

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "'"))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~1k first names with single quotes
-   Most like legitimately quoted first names
-   Some look like they were intended to be diacritics, e.g. ANDRE',
    RENE'

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "'"))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~3k middle names with single quotes
-   Look like legitimately quoted middle names
-   Some look like they were intended to be diacritics, e.g. RENEE'

### Double quote

Check for double quotes.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, '"'))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   1 last name with double quotes

    -   This looks like a typo for a single quote

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, '"'))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   3 first names with double quotes
-   2 are introducing nicknames
-   1 looks like a typo, e.g. HENRYL"

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, '"'))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   1 middle name with double quotes

    -   This this looks like the middle name is the "official" first
        name and the listed first name is a nickname

### Period

Check for periods.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "\\."))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   11 last names with periods
-   Most are the standard contraction of SAINT
-   3 have added the name suffix code JR. to the last name

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "\\."))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   120 first names with periods
-   These look like punctuated initials

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "\\."))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~2k middle names with periods
-   These look like punctuated initials

### Comma

Check for commas.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, ","))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   2 last names with commas
-   Punctuation for suffix field values added to last name

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, ","))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   4 first names with commas
-   Arbitrary added punctuation
-   Punctuation for suffix field value added to first name

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, ","))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   12 middle names with periods
-   List separator
-   Punctuation to squeeze in extra field

### Space

Check for whitespace characters.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "\\s"))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   ~7k last names with whitespace characters
-   Most look like substitutions for hyphen or quote
-   Some are word separators

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "\\s"))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   ~12k first names with whitespace characters
-   Most look like word separators

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "\\s"))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~47k middle names with whitespace characters
-   Most look like word separators

### Other non-alphanumeric

Check for other non-alphanumeric characters.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "[^ a-zA-Z0-9\\.,'\"-]"))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   30 last names with other non-alphanumeric characters
-   Most look like substitutions for hyphen or quote
-   Some look like random cruft

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "[^ a-zA-Z0-9\\.,'\"-]"))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   100 first names with other non-alphanumeric characters
-   Some look like substitutions for hyphen or quote
-   Some are parenthetical notes
-   Some look like random cruft

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "[^ a-zA-Z0-9\\.,'\"-]"))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   \~1k middle names with other non-alphanumeric characters
-   Some look like substitutions for hyphen
-   Many are parenthetical notes (NMN = no middle name)

## Digits

Check for digits.

### Zero

Check for zero

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "0"))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   29 last names with zero
-   Substitution for letter O

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "0"))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   33 first names with zero
-   Substitution for letter O

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "0"))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   77 middle names with zero
-   Substitution for letter O
-   Some are in superfluous numbers

### One

Check for one.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "1"))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   1 last name with one
-   Substitution for I in generation suffix (111 = III)

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "1"))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   0 first names with one

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "1"))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   39 middle names with one
-   Some are substitution for I in generation suffix
-   Some are in superfluous numbers

### Other digits

Check for other digits.

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(last_name, "[2-9]"))

nrow(x)

x %>%   
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(last_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   1 last name with digits 2-9
-   Random insertion

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(first_name, "[2-9]"))

nrow(x)

x %>%   
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(first_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   2 first names with digits 2-9
-   Look like random insertions

```{r}
x <- d %>% 
  dplyr::filter(stringr::str_detect(midl_name, "[2-9]"))

nrow(x)

x %>%   
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(midl_name) %>% 
  dplyr::select(all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   24 middle names with digits 2-9
-   One random insertion
-   Most appear to be superfluous numbers (from the address?)

## Special words

Look for special words that shouldn't be in names.

Define word patterns to search for.

```{r}
# honorifics
w_hons <- c(
  "MR", "MISTER", "MASTER", "MRS", "MS", "MISS", 
  "REV", "REVEREND", "SR", "SISTER", "BR", "BROTHER",
  "FATHER", "MOTHER", "PASTOR", "ELDER", "BISHOP",
  "DR", "DOCTOR", "MD", "PROF", "PROFESSOR"
)

# generation suffixes
w_gen <- c(
  "JR", "JNR", "JUNIOR", "SR", "SNR", "SENIOR",
  "1ST", "2ND", "3RD", "4TH", "5TH", "6TH", "7TH", "8TH",
  "FIRST", "SECOND", "THIRD", "FOURTH", "FIFTH", "SIXTH", "SEVENTH", "EIGHTH", "EIGHTTH",
  "1", "2", "3", "4", "5", "6", "7", "8",
  "I", "II", "III", "IIII", "IV", "V", "VI"
)

# special values
w_spec <- c(
  "NN", "NMN", "NAME",
  "UNK", "UNKNOWN", "AKA", "KNOWN AS", "ALSO KNOWN AS", "ALIAS",
  "BLIND"
)

# test
w_test <- c(
  "TEST", "TST", "DUMMY", "VOTER",  "([A-Z])\\1{2,}"
)
```

### Last name

```{r}
# regular expression to match words
w_regexp <- 
  c(w_hons, w_gen, w_spec, w_test) %>% # all special words
  unique() %>% # make it a set
  dplyr::setdiff( # remove words that appear to mostly be validly used
    c(
      "BISHOP",
      "BLIND",
      "BROTHER",
      "DOCTOR",
      "ELDER",
      "FIRST",
      "JUNIOR",
      "MASTER",
      "MISS",
      "MISTER",
      "NAME",
      "PASTOR",
      "SENIOR",
      "TEST",
      "THIRD",
      "VOTER"
    )
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") # search for any

x <- d %>% 
  dplyr::mutate(
    match = 
      last_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_extract(pattern = w_regexp)
  ) %>% 
  dplyr::filter(!is.na(match))

nrow(x)

x %>% 
  dplyr::arrange(match, sex, last_name, first_name) %>% 
  dplyr::select(match, sex, age, all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

I eyeballed the results and removed words (using `setdiff` in the code above) which appeared to be mostly
validly used.

Invalid words:

-   Most appear to be a suffix code added into the last name

### First name

```{r}
# regular expression to match words
w_regexp <- 
  c(w_hons, w_gen, w_spec, w_test) %>% # all special words
  unique() %>% # make it a set
  dplyr::setdiff( # remove words that appear to mostly be validly used
    c(
      "BISHOP",
      "BROTHER",
      "DOCTOR",
      "ELDER",
      "JUNIOR",
      "MASTER",
      "MISTER",
      "PASTOR",
      "PROFESSOR"
    )
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") # search for any

x <- d %>% 
  dplyr::mutate(
    match = 
      first_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_extract(pattern = w_regexp)
  ) %>% 
  dplyr::filter(!is.na(match))

nrow(x)

x %>% 
  dplyr::arrange(match, sex, last_name, first_name) %>% 
  dplyr::select(match, sex, age, all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

I eyeballed the results and removed words which appeared to be mostly
validly used.

Invalid words:

-   As whole field: FATHER, III, IV, JR, MD, MR, MRS, SISTER, SR
-   As first word: DR, MISS, MRS, REV, SISTER
-   As last word: III, JR, MRS, NMN, SR
-   As internal word: MRS

### Middle name

```{r}
# regular expression to match words
w_regexp <- 
  c(w_hons, w_gen, w_spec, w_test) %>% # all special words
  unique() %>% # make it a set
  dplyr::setdiff( # remove words that appear to mostly be validly used
    c(
      "BISHOP",
      "BLIND",
      "BR",
      "BROTHER",
      "DOCTOR",
      "ELDER",
      "FIRST",
      "JR", # invalid & too many to display 
      "JUNIOR",
      "MASTER",
      "MISTER",
      "MRS", # invalid & too many to display
      "NMN", # invalid & too many to display
      "PASTOR",
      "SENIOR",
      "SISTER",
      "I",
      "V",
      "VI",
      "VOTER"
    )
  ) %>% 
  glue::glue(x = . , "\\b{x}\\b") %>%  # must be words
  glue::glue_collapse(sep = "|") # search for any

x <- d %>% 
  dplyr::mutate(
    match = 
      midl_name %>% 
      stringr::str_to_upper() %>% 
      stringr::str_replace_all(pattern = "[^ A-Z]", replacement = " ") %>% 
      stringr::str_squish() %>% 
      stringr::str_extract(pattern = w_regexp)
  ) %>% 
  dplyr::filter(!is.na(match))

nrow(x)

x %>% 
  dplyr::arrange(match, sex, last_name, first_name) %>% 
  dplyr::select(match, sex, age, all_of(vars_name)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

I eyeballed the results and removed words which appeared to be mostly
validly used.

Invalid words:

-   As whole field: AKA, DR, II, III, IV, JR, MD, MISS, MRS, MS, NMN,
    REV, SR
-   As first word: JR, MRS
-   As last word: DR, II, III, IV, JR, MD, MISS, MR, MRS, NMN, NN, SR
-   As internal word: JR

# Timing {.unnumbered}

```{r echo=FALSE}
tictoc::toc()
```
