---
title: "[meta] Apply row exclusions"
subtitle: "m_01_2_exclusions"
author: "Ross Gayler"
date: "2021-05-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup}
# NOTE this notebook can be run manually or automatically by {targets}
# So load the packages required by this notebook here
# rather than relying on _targets.R to load them.

# Set up the project environment, because {workflowr} knits each Rmd file 
# in a new R session, and doesn't execute the project .Rprofile

library(targets) # access data from the targets cache

library(tictoc) # capture execution time
library(here) # construct file paths relative to project root
library(fs) # file system operations
library(dplyr) # data wrangling
library(gt) # table formatting
library(stringr) # string matching
library(vroom) # fast reading of delimited text files

# start the execution time clock
tictoc::tic("Computation time (excl. render)")

# Get the path to the raw entity data file
# This is a target managed by {targets}
f_entity_raw_tsv <- tar_read(c_raw_entity_data_file)
```

```{r, include = FALSE, cache = FALSE}
# Get all the functions used by {targets} 
# so we can display and use them in this notebook
# with a guarantee that they will be identical
# here and when used by {targets}.

# This is  work-around because tarchetypes::tar_knitr_deps()
# fails if a call to here::here() is embedded in 
# a call to knitr::read_chunk()
knitr::read_chunk("R/functions.R")
```

# Introduction

These meta notebooks document the development of functions that will be
applied in the core pipeline.

The aim of the m_01 set of meta notebooks is to work out how to read the
raw entity data, drop excluded cases, discard irrelevant variables,
apply any cleaning, and construct standardised names. This does not
include construction of any modelling features. To be clear, the target
(`c_raw_entity_data`) corresponding to the objective of this set of
notebooks is the cleaned and standardised raw data, before constructing
any modelling features.

This notebook documents the process of excluding data rows that are not
useful for this project.

The subsequent notebooks in this set will develop the other functions
needed to generate the cleaned and standardised data.

## Exclusions

There are four variables dealing with voter status. Preliminary
examination of these variables shows that some records correspond to
people who have been removed from the electoral roll. This project
focuses on ambiguity arising from the fact that some names are common
Therefore, we want the entity data to be as accurate as possible and
free of duplicate records so that we don't introduce ambiguity because
of data quality issues.

Speaking of data quality issues, it is my experience that large
databases often contain test records.

I will exclude all the data rows that have any of these data quality
issues. I will do this early in the pipeline to minimise the number of
records processed and to avoid including these records in the subsequent
quality analyses.

# Read entity data

Read the raw entity data file using the previously defined core pipeline
function, `raw_entity_data_read()`.

```{r, raw_entity_data_read, echo = FALSE}
```

```{r}
# Show the data file name
fs::path_file(f_entity_raw_tsv)

# Read the raw entity data
d <- raw_entity_data_read(f_entity_raw_tsv)
```

# Voter status variables

Check the internal consistency of the voter status variables.

The data dictionary describes these as two pairs of variables, where
each pair consists of a code variable and a label variable. I expect the
code and label values to be in a 1:1 relationship.

## Status

-   `status_cd` - Status code for voter registration
-   `voter_status_desc` - Status code descriptions

```{r}
d %>% 
  dplyr::distinct(status_cd, voter_status_desc) %>% 
  dplyr::arrange(status_cd, voter_status_desc) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = gt::cell_text(weight = "bold"), locations = gt::cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   `<NA>` is the missing value indicator.
-   The code and label values are in a 1:1 relationship. Use the label
    variable (`voter_status_desc`) because it is more explicitly
    meaningful.

Look at the distribution of values across the data.

```{r}
d %>% 
  dplyr::count(voter_status_desc) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   \~60% of records are `ACTIVE`. All the other values have negative
    connotations for expected data quality.

## Reason

-   `reason_cd` - Reason code for voter registration status
-   `voter_status_reason_desc` - Reason code description

```{r}
d %>% 
  dplyr::distinct(reason_cd, voter_status_reason_desc) %>% 
  dplyr::arrange(reason_cd, voter_status_reason_desc) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = gt::cell_text(weight = "bold"), locations = gt::cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   `<NA>` is the missing value indicator.
-   The code and label values are in a 1:1 relationship. Use the label
    variable (`voter_status_reason_desc`) because it is more explicitly
    meaningful.

Look at the distribution of values.

```{r}
d %>% 
  dplyr::count(voter_status_reason_desc) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   `<NA>` is the missing value indicator.
-   \~50% of records are "VERIFIED". That value seems to have the most
    positive connotations for expected data quality.

## Status $\times$ Reason

Look at the distribution of combinations of the two variables.

```{r}
d %>% 
  dplyr::count(voter_status_desc, voter_status_reason_desc) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   \~50% are "ACTIVE" and "VERIFIED"

On a common-sense interpretation of the labels, these are the records
that have survived the registration checking process, so are most likely
free of errors and duplicates.

Write a function to filter the records to keep only those that are
"ACTIVE" and "VERIFIED".

```{r, raw_entity_data_excl_status}
```

Apply the filter before moving on to the next exclusion condition and
track the number of rows before and after the filter.

```{r}
# number of rows before filtering
nrow(d)

d <- d %>% raw_entity_data_excl_status()

# number of rows after filtering
nrow(d)
```

-   Result as expected

# Test cases

Initial poking through the data showed the words DUMMY, PRACTICE, TEST,
THIS, THISIS, and THISISA as person names appeared to be associated with
records that were subjectively judged as being likely to be test cases.
Look for any records that contain those strings as a word (i.e. enclosed
by word boundaries) in the person name columns.

```{r}
# define the target regular expression
target <- regex(
  "\\bDUMMY\\b|\\bPRACTICE\\b|\\bTEST\\b|\\bTHIS\\b|\\bTHISIS\\b|\\bTHISISA\\b", 
  ignore_case = TRUE
)

d %>% 
  dplyr::select(
    last_name, first_name, midl_name,
    street_name, street_type_cd, res_city_desc,
    sex, phone_num
  ) %>% 
  dplyr::filter(
    stringr::str_detect(last_name, target)|
      stringr::str_detect(first_name, target) |
      stringr::str_detect(midl_name, target)
  ) %>% 
  dplyr::arrange(last_name, street_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::tab_style(
    style = gt::cell_fill(color = "yellow"),
    locations = list(
      gt::cells_body(columns = last_name, rows = stringr::str_detect(last_name, target)),
      gt::cells_body(columns = first_name, rows = stringr::str_detect(first_name, target)),
      gt::cells_body(columns = midl_name, rows = stringr::str_detect(midl_name, target))
    )
  )
```

-   Online search finds that "Test" is a known last name in the USA.
-   The name "TEST, THIS" looks like a test case (with ACTIVE VERIFIED
    status).

Write a function to filter the records to exclude those with name "TEST,
THIS".

There are likely to be many other test cases not detected by this
filter. However, they are probably only a tiny fraction of the records -
so it's not a big problem for this project if they are missed. If I find
any other test cases later I will come back here and revise the
exclusion criteria.

```{r, raw_entity_data_excl_test}
```

Apply the filter and track the number of rows before and after the
filter.

```{r}
# number of rows before filtering
nrow(d)

d <- d %>% raw_entity_data_excl_test()

# number of rows after filtering
nrow(d)
```

-   Behaviour as expected

# Timing {.unnumbered}

```{r echo=FALSE}
tictoc::toc()
```
