---
title: "[meta] Clean all the variables"
subtitle: "m_01_9_clean_vars"
author: "Ross Gayler"
date: "2021-05-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup}
# NOTE this notebook can be run manually or automatically by {targets}
# So load the packages required by this notebook here
# rather than relying on _targets.R to load them.

# Set up the project environment, because {workflowr} knits each Rmd file 
# in a new R session, and doesn't execute the project .Rprofile

library(targets) # access data from the targets cache

library(tictoc) # capture execution time
library(here) # construct file paths relative to project root
library(fs) # file system operations
library(dplyr) # data wrangling
library(gt) # table formatting
library(vroom) # fast reading of delimited text files
library(tibble) # enhanced data frames
library(stringr) # string matching
library(skimr) # compact summary of each variable
library(lubridate) # date parsing
library(forcats) # manipulation of factors
library(ggplot2) # graphics
library(tidyr) # data tidying
#

# start the execution time clock
tictoc::tic("Computation time (excl. render)")

# Get the path to the raw entity data file
# This is a target managed by {targets}
f_entity_raw_tsv <- tar_read(c_raw_entity_data_file)
```

```{r, include = FALSE, cache = FALSE}
# Get all the functions used by {targets} 
# so we can display and use them in this notebook
# with a guarantee that they will be identical
# here and when used by {targets}.

# This is  work-around because tarchetypes::tar_knitr_deps()
# fails if a call to here::here() is embedded in 
# a call to knitr::read_chunk()
knitr::read_chunk("R/functions.R")
```

# Introduction

The aim of this set of meta notebooks is to work out how to read the raw
entity data. and get it sufficiently neatened so that we can construct
standardised names and modelling features without needing any further
neatening. To be clear, the target (`c_raw_entity_data`) corresponding
to the objective of this set of notebooks is the neatened raw data,
before constructing any modelling features.

This notebook documents the cleaning of all the variables based on the
issues discovered in the previous analyses.

It also adds a record ID variable.

## Variable roles

The name variables, `last_name`, `first_name`, and `midl_name`, will
definitely be used in compatibility modelling.

We intend to use the one snapshot file as both the database to be
queried and as the set of queries. Consequently, strictly speaking, we
don't need to standardise the name variables because the database and
query records are guaranteed to be identical (they will literally be the
same record). However, we will look at the name variables with an eye to
standardisation because it is never a good idea to statistically model
data without having an idea about the quality of the data. We will apply
some basic standardisation to the name variables, if appropriate,
because it parallels what would be necessary in practice.

The demographic variables `sex`, `age`, `birth_place`, and the
administrative variable `county_desc` may be used as predictors and/or
blocking variables.

The remainder of the variables (residence and administrative variables)
will be kept in case they are useful for manually assessing claimed
matches.

## Cleanup for predictors

### Name standardisation

Standardisation will be applied to the name variables `last_name`,
`first_name`, and `midl_name`. This attempts to remove variation that is
probably irrelevant to identity (e.g. case, punctuation, and spacing).

### Missing values

In the previous notebooks I have converted empty strings to missing
values (`NA_character_` in R). This was convenient because `table()` and
`skim()` count missing values as a separate category. However, modelling
is a different kettle of fish.

In modelling, we want to get an estimated probability of identity match
for every query, regardless of how many attributes have missing values.
Typical modelling functions do not tolerate any missing (`NA`) values in
predictors. If any of the predictors is missing then the estimate is
also missing.

We avoid that problem by transforming the missing values into some
nonmissing value and creating an extra variable to indicate the
missingness. This will be done for the name variables `last_name`,
`first_name`, and `midl_name`, and the demographic variable `age`. (This
is not necessary for `birth_place` because "missing" is just another
valid level of the variable.

### Cleanup summary

The cleanup actions to be applied are (in order):

-   *age*

-   convert from string to integer

-   add missing value indicator and set to true if age \< 17 or age \>
    104

-   if age missing indicator is true, set age to 0

-   *preprocess all character variables*

-   map missing to empty string

-   map lower case letters to upper case

-   *all names*

-   map each non-alphanumeric character to a space (Remove variability
    of punctuation while preserving word boundaries.)

-   map words 11, 111, 1111 to words II, III, IIII (Correct substitution
    of 1 for I in generation suffixes.)

-   if name contains zero and no other digits, map zero to O (Correct
    substitution of 0 for O in names.)

-   map each remaining digit to an empty string (Remove random digit
    insertions)

-   *last name*

-   map words DR, II, III, IIII, IV, JR, MD, SR to empty string

-   if number of letters in last name = 1, map name to empty string

-   *middle name*

-   map words AKA, DR, II, III, IV, JR, MD, MISS, MR, MRS, MS, NMN, NN,
    REV, SR to empty string

-   *first name*

-   map words DR, FATHER, III, IV, JR, MD, MISS, MR, MRS, NMN, REV,
    SISTER, SR to empty string

-   if number of letters in first name = 0, move first word of middle
    name to first name

-   *postprocess all name variables*

-   map all spaces to empty strings (Remove variability of spacing.)

-   add missing value indicator variables for all name variables

# Read entity data

Read the raw entity data file using the previously defined functions
`raw_entity_data_read()`, `raw_entity_data_excl_status()`,
`raw_entity_data_excl_test()`, `raw_entity_data_drop_novar()`,
`raw_entity_data_parse_dates()`, and `raw_entity_data_drop_cancel_dt()`.

```{r, raw_entity_data_read, echo = FALSE}
```

```{r, raw_entity_data_excl_status, echo = FALSE}
```

```{r, raw_entity_data_excl_test, echo = FALSE}
```

```{r, raw_entity_data_drop_novar, echo = FALSE}
```

```{r, raw_entity_data_parse_dates, echo = FALSE}
```

```{r, raw_entity_data_drop_admin, echo = FALSE}
```

```{r, raw_entity_data_drop_demog, echo = FALSE}
```

```{r}
# Show the data file name
fs::path_file(f_entity_raw_tsv)

d <- raw_entity_data_read(f_entity_raw_tsv) %>% 
  raw_entity_data_excl_status() %>% 
  raw_entity_data_excl_test() %>% 
  raw_entity_data_drop_novar() %>% 
  raw_entity_data_parse_dates() %>% 
  raw_entity_data_drop_admin() %>% 
  raw_entity_data_drop_demog()

dim(d)
```

# Define and apply cleanup functions

### Age

-   convert from string to integer
-   add missing indicator and set to true if age \< 17 or age \> 104
-   if age missing indicator is true, set age to 0

```{r, raw_entity_data_clean_age}
```

Apply and check function.

```{r}
# Apply

dim(d)

d <- d %>% raw_entity_data_clean_age()

dim(d)

# After

summary(d$age_cln)

summary(d$age_cln_miss)

d %>% 
  dplyr::select(starts_with("age_cln")) %>% 
  dplyr::group_by(age_cln_miss) %>% 
  dplyr::summarise(qtl = quantile(age_cln) %>% list()) %>%
  tidyr::unnest_wider(col = qtl) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels())
```

### Preprocess all character variables

-   map missing to empty string
-   map lower case letters to upper case

Note that this is an in-place transformation of the character variables
rather than adding the transformed values as new variables. This is
because I am viewing this a light tidying rather than as creating
distinctly new cleaned values.

```{r, raw_entity_data_clean_preprocess_char}
```

Apply and check function.

```{r}
# This is only a partial check - look at a very small subset of variables

# Before

d %>% 
  dplyr::select(ends_with("_name")) %>% 
  skimr::skim()

d %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::filter(stringr::str_detect(first_name, "[a-z]")) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

# Apply

d <- d %>% raw_entity_data_clean_preprocess_char()

# After

d %>% 
  dplyr::select(ends_with("_name")) %>% 
  skimr::skim()

d %>% 
  dplyr::select(ends_with("_name")) %>% 
  dplyr::filter(stringr::str_detect(first_name, "[a-z]")) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

### All names

-   map each non-alphanumeric character to a space (Remove variability
    of punctuation. This preserves word boundaries.)
-   map words 11, 111, 1111 to words II, III, IIII (Correct substitution
    of 1 for I in generation suffixes.)
-   if name contains zero and no other digits, map zero to O (Correct
    substitution of 0 for O in names.)
-   map each digit to an empty string (Remove random digit insertions)

```{r, raw_entity_data_clean_all_names}
```

Apply and check function.

```{r}
# This is only a partial check

# Apply

dim(d)

d <- d %>% raw_entity_data_clean_all_names()

dim(d)

# After

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(first_name, "[^ a-zA-Z0-9]")) %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(midl_name, "111*")) %>% 
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(first_name, "0")) %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(midl_name, "[2-9]")) %>% 
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

### Last name

-   map words DR, II, III, IIII, IV, JR, MD, SR to empty string
-   if number of letters in last name = 1, map name to empty string

```{r, raw_entity_data_clean_last_name}
```

Apply and check function.

```{r}
# This is only a partial check

# Before

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(
    last_name_cln, 
    "\\bDR\\b|\\bII\\b|\\bIII\\b|\\bIIII\\b|\\bIV\\b|\\bJR\b|\\bMD\\b|\\bSR\\b")
  ) %>% 
  dplyr::distinct(last_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_length(last_name_cln) <= 1) %>% 
  dplyr::distinct(last_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

# Apply

dim(d)

d <- d %>% raw_entity_data_clean_last_name()

dim(d)

# After

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(
    last_name_cln, 
    "\\bDR\\b|\\bII\\b|\\bIII\\b|\\bIIII\\b|\\bIV\\b|\\bJR\b|\\bMD\\b|\\bSR\\b")
  ) %>% 
  dplyr::distinct(last_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_length(last_name_cln) <= 1) %>% 
  dplyr::distinct(last_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

### Middle name

-   map words AKA, DR, II, III, IV, JR, MD, MISS, MR, MRS, MS, NMN, NN,
    REV, SR to empty string

```{r, raw_entity_data_clean_middle_name}
```

Apply and check function.

```{r}
# This is only a partial check

# Before

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(
    midl_name_cln, 
    "\\bAKA\\b|\\bDR\\b|\\bII\\b|\\bIII\\b|\\bIV\\b|\\bJR\b|\\bMD\\b|\\bMISS\\b|\\bMR\\b|\\MRS\\b|\\bMS\\b|\\bNMN\\b|\\bNN\\b|\\bREV\\b|\\bSR\\b")
  ) %>% 
  dplyr::distinct(midl_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

# Apply

dim(d)

d <- d %>% raw_entity_data_clean_middle_name()

dim(d)

# After

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(
    midl_name_cln, 
    "\\bAKA\\b|\\bDR\\b|\\bII\\b|\\bIII\\b|\\bIV\\b|\\bJR\b|\\bMD\\b|\\bMISS\\b|\\bMR\\b|\\MRS\\b|\\bMS\\b|\\bNMN\\b|\\bNN\\b|\\bREV\\b|\\bSR\\b")
  ) %>% 
  dplyr::distinct(midl_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

### First name

-   map words DR, FATHER, III, IV, JR, MD, MISS, MR, MRS, NMN, REV,
    SISTER, SR to empty string
-   if number of letters in first name = 0, move first word of middle
    name to first name

```{r, raw_entity_data_clean_first_name}
```

Apply and check function.

```{r}
# This is only a partial check

# Before

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(
    first_name_cln, 
    "\\bDR\\b|\\bFATHER\\b|\\bIII\\b|\\bIV\\b|\\bJR\b|\\bMD\\b|\\bMISS\\b|\\bMR\\b|\\MRS\\b|\\bNMN\\b|\\bREV\\b|\\bSISTER\\b|\\bSR\\b")
  ) %>% 
  dplyr::distinct(first_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_length(first_name_cln) == 0) %>% 
  dplyr::distinct(first_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

# Apply

dim(d)

d <- d %>% raw_entity_data_clean_first_name()

dim(d)

# After

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(
    first_name_cln, 
    "\\bDR\\b|\\bFATHER\\b|\\bIII\\b|\\bIV\\b|\\bJR\b|\\bMD\\b|\\bMISS\\b|\\bMR\\b|\\MRS\\b|\\bNMN\\b|\\bREV\\b|\\bSISTER\\b|\\bSR\\b")
  ) %>% 
  dplyr::distinct(first_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_length(first_name_cln) == 0) %>% 
  dplyr::distinct(first_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

### Postprocess all name variables

-   map all spaces to empty strings (Remove variability of spacing.)
-   add missing value indicator variables for all name variables

```{r, raw_entity_data_clean_postprocess_names}
```

Apply and check function.

```{r}
# This is only a partial check

# Before

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(first_name_cln, " ")) %>% 
  dplyr::distinct(first_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

# Apply

dim(d)

d <- d %>% raw_entity_data_clean_postprocess_names()

dim(d)

# After

d %>% 
  dplyr::select(
    last_name, last_name_cln, first_name, first_name_cln, midl_name, midl_name_cln
  ) %>% 
  dplyr::filter(stringr::str_detect(first_name_cln, " ")) %>% 
  dplyr::distinct(first_name_cln, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

# Examples

Define a function to apply all the name cleaning actions.

```{r, raw_entity_data_clean_all}
```

This function is equivalent to the incremental processing that occurred
in the earlier chunks.

Show some examples of the cleaned data.

Quick distributions

```{r}
d %>% 
  dplyr::select(ends_with("cln"), ends_with("_miss")) %>% 
  skimr::skim()
```

## Age

```{r}
d %>% 
  dplyr::select(age, age_cln, age_cln_miss) %>% 
  dplyr::group_by(age_cln_miss) %>% 
  dplyr::distinct(age, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 10) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

## Last name

```{r}
d %>% 
  dplyr::select(
    last_name, last_name_cln, last_name_cln_miss,
    first_name, first_name_cln, first_name_cln_miss,
    midl_name, midl_name_cln, midl_name_cln_miss
  ) %>% 
  dplyr::group_by(last_name_cln_miss) %>% 
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 10) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

```{r}
d %>% 
  dplyr::select(
    last_name, last_name_cln, last_name_cln_miss,
    first_name, first_name_cln, first_name_cln_miss,
    midl_name, midl_name_cln, midl_name_cln_miss
  ) %>% 
  dplyr::filter(last_name != last_name_cln) %>% 
  dplyr::distinct(last_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

## First name

```{r}
d %>% 
  dplyr::select(
    last_name, last_name_cln, last_name_cln_miss,
    first_name, first_name_cln, first_name_cln_miss,
    midl_name, midl_name_cln, midl_name_cln_miss
  ) %>% 
  dplyr::group_by(first_name_cln_miss) %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 10) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

```{r}
d %>% 
  dplyr::select(
    last_name, last_name_cln, last_name_cln_miss,
    first_name, first_name_cln, first_name_cln_miss,
    midl_name, midl_name_cln, midl_name_cln_miss
  ) %>% 
  dplyr::filter(first_name != first_name_cln) %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

```{r}
d %>% 
  dplyr::select(
    last_name, last_name_cln, last_name_cln_miss,
    first_name, first_name_cln, first_name_cln_miss,
    midl_name, midl_name_cln, midl_name_cln_miss
  ) %>% 
  dplyr::filter(stringr::str_detect(first_name, "SISTER")) %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

```{r}
d %>% 
  dplyr::select(
    last_name, last_name_cln, last_name_cln_miss,
    first_name, first_name_cln, first_name_cln_miss,
    midl_name, midl_name_cln, midl_name_cln_miss
  ) %>% 
  dplyr::filter(first_name == "") %>% 
  dplyr::distinct(first_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

## Middle name

```{r}
d %>% 
  dplyr::select(
    last_name, last_name_cln, last_name_cln_miss,
    first_name, first_name_cln, first_name_cln_miss,
    midl_name, midl_name_cln, midl_name_cln_miss
  ) %>% 
  dplyr::group_by(midl_name_cln_miss) %>% 
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 10) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

```{r}
d %>% 
  dplyr::select(
    last_name, last_name_cln, last_name_cln_miss,
    first_name, first_name_cln, first_name_cln_miss,
    midl_name, midl_name_cln, midl_name_cln_miss
  ) %>% 
  dplyr::filter(midl_name != midl_name_cln) %>% 
  dplyr::distinct(midl_name, .keep_all = TRUE) %>% 
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

# Putting it all together

Define a function to add a record ID variable.

```{r, raw_entity_data_add_id}
```

Define a function to apply all the steps from reading the raw entity
data through to creating the cleaned entity data. This will be used in
the core pipeline.

```{r, raw_entity_data_make_clean}
```

# Timing {.unnumbered}

```{r echo=FALSE}
tictoc::toc()
```
