---
title: "[meta] Check residenial variables"
subtitle: "m_01_6_check_resid"
author: "Ross Gayler"
date: "2021-05-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup}
# NOTE this notebook can be run manually or automatically by {targets}
# So load the packages required by this notebook here
# rather than relying on _targets.R to load them.

# Set up the project environment, because {workflowr} knits each Rmd file 
# in a new R session, and doesn't execute the project .Rprofile

library(targets) # access data from the targets cache

library(tictoc) # capture execution time
library(here) # construct file paths relative to project root
library(fs) # file system operations
library(dplyr) # data wrangling
library(gt) # table formatting
library(stringr) # string matching
library(vroom) # fast reading of delimited text files
library(lubridate) # date parsing
library(forcats) # manipulation of factors
library(ggplot2) # graphics
library(skimr) # compact summary of each variable
library(tidyr) # data tidying

# start the execution time clock
tictoc::tic("Computation time (excl. render)")

# Get the path to the raw entity data file
# This is a target managed by {targets}
f_entity_raw_tsv <- tar_read(c_raw_entity_data_file)
```

```{r, include = FALSE, cache = FALSE}
# Get all the functions used by {targets} 
# so we can display and use them in this notebook
# with a guarantee that they will be identical
# here and when used by {targets}.

# This is  work-around because tarchetypes::tar_knitr_deps()
# fails if a call to here::here() is embedded in 
# a call to knitr::read_chunk()
knitr::read_chunk("R/functions.R")
```

# Introduction

These meta notebooks document the development of functions that will be
applied in the core pipeline.

The aim of the m_01 set of meta notebooks is to work out how to read the
raw entity data, drop excluded cases, discard irrelevant variables,
apply any cleaning, and construct standardised names. This does not
include construction of any modelling features. To be clear, the target
(`c_raw_entity_data`) corresponding to the objective of this set of
notebooks is the cleaned and standardised raw data, before constructing
any modelling features.

This notebook documents the checking of the "residential" variables for
any issues that need fixing. These are the residential address and the
phone number (which is tied to the address if the telephone is a
land-line).

The subsequent notebooks in this set will develop the other functions
needed to generate the cleaned and standardised data.

Regardless of whether there are any issues that need to be fixed, the
analyses here may inform our use of these variables in later analyses.

We have no intention of using the residence variables as predictors for
entity resolution. However, they may be of use for manually checking the
results of entity resolution. Consequently, the checking done here is
minimal.

Define the residential variables:

-   `unit_num` - Residential address unit number
-   `house_num` - Residential address street number
-   `half_code` - Residential address street number half code
-   `street_dir` - Residential address street direction (N,S,E,W,NE,SW,
    etc.)
-   `street_name` - Residential address street name
-   `street_type_cd` - Residential address street type (RD, ST, DR,
    BLVD, etc.)
-   `street_sufx_cd` - Residential address street suffix (BUS, EXT, and
    directional)
-   `res_city_desc` - Residential address city name
-   `state_cd` - Residential address state code
-   `zip_code` - Residential address zip code
-   `area_cd` - Area code for phone number
-   `phone_num` - Telephone number

```{r}
vars_resid <- c(
  "unit_num", "house_num",      
  "half_code", "street_dir", "street_name", "street_type_cd", "street_sufx_cd",
  "res_city_desc", "state_cd", "zip_code",
  "area_cd", "phone_num"
)
```

# Read entity data

Read the raw entity data file using the previously defined core pipeline
functions `raw_entity_data_read()`, `raw_entity_data_excl_status()`,
`raw_entity_data_excl_test()`, `raw_entity_data_drop_novar()`,
`raw_entity_data_parse_dates()`, and `raw_entity_data_drop_cancel_dt()`.

```{r, raw_entity_data_read, echo = FALSE}
```

```{r, raw_entity_data_excl_status, echo = FALSE}
```

```{r, raw_entity_data_excl_test, echo = FALSE}
```

```{r, raw_entity_data_drop_novar, echo = FALSE}
```

```{r, raw_entity_data_parse_dates, echo = FALSE}
```

```{r, raw_entity_data_drop_admin, echo = FALSE}
```

```{r}
# Show the data file name
fs::path_file(f_entity_raw_tsv)

d <- raw_entity_data_read(f_entity_raw_tsv) %>% 
  raw_entity_data_excl_status() %>% 
  raw_entity_data_excl_test() %>% 
  raw_entity_data_drop_novar() %>% 
  raw_entity_data_parse_dates() %>% 
  raw_entity_data_drop_admin()
dim(d)
```

# Dwelling

-   `unit_num` - Residential address unit number
-   `house_num` - Residential address street number
-   `half_code` - Residential address street number half code

```{r}
d %>% 
  dplyr::select(unit_num, house_num, half_code) %>% 
  skimr::skim()
```

We are mostly interested in how much these fields are used, so
concentrate on `complete_rate`.

All these variables are character variables, so `min` and `max` refer to
the minimum and maximum lengths of the values as character strings.

The number of unique values, `n_unique`, is also of interest.

-   `unit_num`

    -   8% filled
    -   There are some awfully long unit numbers
    -   There are an awful lot of unique unit numbers

-   `house_num`

    -   100% filled
    -   There are some awfully long house numbers
    -   There are an awful lot of unique house numbers

-   `half_code` 0.3% filled

    -   All exactly 1 character long
    -   There are more unique values than I would expect for a one
        character string

## unit_num

-   `unit_num` - Residential address unit number

Look at some examples grouped by length

```{r}
d %>% 
  dplyr::select(unit_num) %>% 
  dplyr::filter(!is.na(unit_num)) %>% 
  dplyr::mutate(length = stringr::str_length(unit_num)) %>% 
  dplyr::group_by(length) %>% 
  dplyr::count(unit_num) %>% # count occurrences of each unique value
  dplyr::slice_max(order_by = n, n = 5) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   Longer values are due to inclusion of text, e.g. "APT-106"

## house_num

-   `house_num` - Residential address street number

Look at some examples grouped by length

```{r}
d %>% 
  dplyr::filter(!is.na(house_num)) %>% 
  dplyr::count(house_num) %>% # count occurrences of each unique value
  dplyr::mutate(length = stringr::str_length(house_num)) %>% 
  dplyr::group_by(length) %>% 
  dplyr::slice_max(order_by = n, n = 5, with_ties = FALSE) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   I am mildly surprised by house number "0". I wouldn't be surprised
    if someone was using that as a missing value flag.
-   Large numbers are plausible, because these are not uncommon in the
    USA.
-   Very large numbers are somewhat suspect.

## half_code

-   `half_code` - Residential address street number half code

```{r}
d %>% 
  dplyr::select(half_code) %>% 
  dplyr::filter(!is.na(half_code)) %>% 
  dplyr::count(half_code) %>% # count occurrences of each unique value
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   `half_code` appears to indicate where there are multiple dwellings
    on one street-numbered block. Typical values would be A, B, ...
-   Non-alphanumeric characters are not plausible

# Street

-   `street_dir` - Residential address street direction (N,S,E,W,NE,SW,
    etc.)
-   `street_name` - Residential address street name
-   `street_type_cd` - Residential address street type (RD, ST, DR,
    BLVD, etc.)
-   `street_sufx_cd` - Residential address street suffix (BUS, EXT, and
    directional)

Take a quick look at the summaries.

```{r}
d %>% 
  dplyr::select(starts_with("street_")) %>% 
  skimr::skim()
```

We are mostly interested in how much these fields are used, so
concentrate on `complete_rate`.

All these variables are character variables, so `min` and `max` refer to
the minimum and maximum lengths of the values as character strings.

The number of unique values, `n_unique`, is also of interest.

-   `street_dir`

    -   7% filled
    -   1 or 2 characters long
    -   8 unique values

-   `street_name`

    -   \~100% filled (7 missing)
    -   Some very short names
    -   Wide range of lengths
    -   Many unique values

-   `street_type_cd`

    -   96% filled
    -   2 to 4 characters long
    -   119 unique values

-   `street_sufx_cd`

    -   4% filled
    -   1 to 3 characters long
    -   11 unique values

## street_dir

-   `street_dir` - Residential address street direction (N,S,E,W,NE,SW,
    etc.)

Look at the distribution of values.

```{r}
d %>% 
  dplyr::count(street_dir) %>% 
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   They are literally just compass directions.

## street_name

-   `street_name` - Residential address street name

Seven records are missing street name. Look at them.

```{r}
d %>% 
  dplyr::filter(is.na(street_name)) %>% 
  dplyr::select(unit_num, house_num, half_code, starts_with("street_"), res_city_desc:zip_code) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   Seven records have no residential address. Are these homeless people
    or data entry errors?

Look at some other information for the same records.

```{r}
d %>% 
  dplyr::filter(is.na(street_name)) %>% 
  dplyr::select(last_name, first_name, sex, age, birth_place, phone_num) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   Lots of missing values for these people.

Some street names are very short. Look at the distribution of length of
street name.

```{r}
summary(stringr::str_length(d$street_name))

d %>% 
  ggplot() +
  geom_histogram(aes(x = stringr::str_length(street_name) %>% replace_na(0)), binwidth = 1) +
  scale_y_sqrt()
```

Look at examples of short street names.

```{r}
# length = 1
d %>% 
  dplyr::filter(stringr::str_length(street_name) == 1) %>% 
  dplyr::select(starts_with("street_"), res_city_desc) %>%
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")

# length = 2
d %>% 
  dplyr::filter(stringr::str_length(street_name) == 2) %>% 
  dplyr::select(starts_with("street_"), res_city_desc) %>%
  dplyr::slice_sample(n = 20) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

I checked some of these examples against a map.

-   Some streets have names like A, B, C, ...
-   "JC Road" is a valid street name

Look at examples of long street names.

```{r}
# length >= 28
d %>% 
  dplyr::filter(stringr::str_length(street_name) >= 28) %>% 
  dplyr::select(starts_with("street_"), res_city_desc) %>%
  dplyr::distinct(.keep_all = TRUE) %>% 
  dplyr::arrange(street_name) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   Long street names are multi-word phrases

    -   Some appear to have been truncated

## street_type_cd

-   `street_type_cd` - Residential address street type (RD, ST, DR,
    BLVD, etc.)

```{r}
d %>% 
  dplyr::count(street_type_cd) %>% 
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

## street_sufx_cd

-   `street_sufx_cd` - Residential address street suffix (BUS, EXT, and
    directional)

```{r}
d %>% 
  dplyr::count(street_sufx_cd) %>% 
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

# Locality

-   `res_city_desc` - Residential address city name\
-   `state_cd` - Residential address state code\
-   `zip_code` - Residential address zip code

```{r}
d %>% 
  dplyr::select(res_city_desc : zip_code) %>% 
  skimr::skim()
```

-   `res_city_desc` \~100% filled (19 missing)
-   `state_cd` \~100% filled (18 missing)
-   `zip_code` \~100% filled (21 missing)

Look at the addresses with any missing locality variable.

```{r}
d %>% 
  dplyr::filter(is.na(res_city_desc) | is.na(state_cd) | is.na(zip_code)) %>% 
  dplyr::select(house_num : zip_code) %>%
  dplyr::distinct(.keep_all = TRUE) %>%
  dplyr::arrange(state_cd, res_city_desc, zip_code) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   Some appear to be good addresses, apart from a missing zip code
-   One appears to be a good address, apart from a missing state code
-   Some are CONFIDENTIAL addresses, with all details missing
-   Some appear to be completely missing the address (homeless persons?)

## res_city_desc

-   `res_city_desc` Residential address city name\

Some city names are very short. Look at the distribution of length of
city name.

```{r}
d %>% pull(res_city_desc) %>% stringr::str_length() %>% summary()

d %>% 
  ggplot() +
  geom_histogram(aes(x = stringr::str_length(res_city_desc) %>% replace_na(0)), binwidth = 1) +
  scale_y_sqrt()
```

Look at examples of short city names.

```{r}
# length = 3
d %>% 
  dplyr::filter(stringr::str_length(res_city_desc) == 3) %>% 
  dplyr::count(res_city_desc) %>%
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)

# length = 4
d %>% 
  dplyr::filter(stringr::str_length(res_city_desc) == 4) %>% 
  dplyr::count(res_city_desc) %>%
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   These look like plausible place names

Look at examples of long city names.

```{r}
# length >= 17
d %>% 
  dplyr::filter(stringr::str_length(res_city_desc) >= 17) %>% 
  dplyr::count(res_city_desc) %>%
  dplyr::slice_sample(n = 20) %>% 
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   Long city names are multi-word phrases

## state_cd

-   `state_cd` - Residential address state code

```{r}
d %>% 
  dplyr::count(state_cd) %>% 
  dplyr::arrange(desc(n)) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   Residential state codes are almost entirely NC (North Carolina)

    -   There are a small number in neighbouring states (non-resident
        voters?)

## zip_code

`zip_code` Residential address zip code

The zip codes are not all the same length. Look at the distribution of
length of zip code

```{r}
d %>% 
  dplyr::select(zip_code) %>% 
  dplyr::mutate(length = stringr::str_length(zip_code)) %>% 
  dplyr::count(length) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

Look at the 9-digit zip codes.

```{r}
d %>% 
  dplyr::filter(stringr::str_length(zip_code) == 9) %>% 
  dplyr::select(street_name : zip_code) %>%
  dplyr::arrange(zip_code) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>")
```

-   The addresses with 9-digit zipcodes look plausible
-   5-digit and 9-digit zip codes are
    [valid](https://en.wikipedia.org/wiki/ZIP_Code)\

# Telephone

-   `area_cd` - Area code for phone number
-   `phone_num` - Telephone number

```{r}
d %>% 
  dplyr::select(area_cd, phone_num) %>% 
  skimr::skim()
```

- Area code ~36% filled
- Phone number ~38% filled

Look at the relationship between missing area code and missing phone
number.

```{r}
table(area_miss = is.na(d$area_cd), phone_miss = is.na(d$phone_num))
```

-   Area code and phone number are generally missing together, but ...
-   Area code is frequently missing even when phone number is present (possibly assumed to be a local number)
-   Phone number is less frequently missing when area code is present (possibly assumed to be in the local area, but phone number unknown)

## area_cd

-   `area_cd` - Area code for phone number

Look at some area code examples grouped by length, *where there is a
phone number*.

```{r}
d %>% 
  dplyr::filter(!is.na(phone_num)) %>%
  dplyr::count(area_cd) %>% # count occurrences of each unique value
  dplyr::mutate(length = stringr::str_length(area_cd) %>% replace_na(0)) %>% 
  dplyr::group_by(length) %>% 
  dplyr::slice_max(order_by = n, n = 10, with_ties = FALSE) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   Many records with phone numbers are missing area code.
-   Area codes less than 3 characters long are probably typos.
-   Area code "999" and "000" are probably missing.

## phone_num

-   `phone_num` - Telephone number

Look at some phone number examples grouped by length, *where there is a
phone number*.

```{r}
d %>% 
  dplyr::filter(!is.na(phone_num)) %>%
  dplyr::count(phone_num) %>% # count occurrences of each unique value
  dplyr::mutate(length = stringr::str_length(phone_num) %>% replace_na(0)) %>% 
  dplyr::group_by(length) %>% 
  dplyr::slice_max(order_by = n, n = 5, with_ties = FALSE) %>% 
  gt::gt() %>% 
  gt::opt_row_striping() %>% 
  gt::tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) %>% 
  gt::fmt_missing(columns = everything(), missing_text = "<NA>") %>% 
  gt::fmt_number(columns = n, decimals = 0)
```

-   Phone numbers less than 7 characters long are probably typos.
-   Phone numbers "0000000" and "9999999" are probably missing.

# Timing {.unnumbered}

```{r echo=FALSE}
tictoc::toc()
```
